### 2015-03-17 infoflow.tex
infoflow

\chapter{Information-flow Security}

As discussed earlier, in Chapter \ref{XXX}, two of the main threats addressed by application security analysis are \emph{integrity vulnerabilities}, whereby untrusted data is able to reach security-sensitive operations without sufficient validation or sanitization, and \emph{confidentiality vulnerabilities}, whereby sensitive data is released to untrusted observers without sufficient anonymization (or declassification). Pleasingly, while these threats are dual to each other, they both reduce to the same formal problem statement.

\section{Problem Statement}

Consider a representation of the program as a graph, where the nodes are control locations and the edges denote control transitions. For the code example in Figure \ref{Fi:chIFS:example}, for instance, we obtain the graph in Figure \ref{Fi:chIFS:exampleGraph}. For conciseness, we represent control locations in the graph as pairs $(x,y)$, where the first (second) component $x$ ($y$) is the line number of the incoming (outgoing) statement. {\tt EN} and {\tt EX} are privileged entry and exit locations, respectively.

\begin{figure*}
	\begin{minipage}[b]{0.5\textwidth}
		\begin{lstlisting}[numbers=left,basicstyle=\sffamily\small,language=Java,stepnumber=1]
b = nondet();
if (b) {
  // integrity source
  String input = getParameter(); 
  // integrity sink
  renderToHTML(input);
} else {
  // confidentiality source 
  String pwd = readPassword();
  // confidentiality sink
  Log.info("Password: "+pwd);
}
		\end{lstlisting}	
		\caption{\label{Fi:chIFS:example}Program in Java with integrity and confidentiality vulnerabilities}
	\end{minipage}
	\hspace{10pt}
	\begin{minipage}[b]{0.5\textwidth}
\xymatrix@C=0pt@R=20pt{
	& ({\tt EN})\ar[d]^{\tt nondet} & \\
	& (1,2)\ar[dl]_{\tt true}\ar[dr]^{\tt false}  \\
	(2,4)\ar[d]_{\tt getParameter} & & (2,9)\ar[d]^{\tt readPassword} \\
	(4,6)\ar[dr]_{\tt renderToHTML} & & (9,11)\ar[dl]^{\tt info} & \\
	& ({\tt EX}) &
}
		\caption{\label{Fi:chIFS:exampleGraph}The control-flow graph corresponding to the code in Figure \ref{Fi:chIFS:example}}
	\end{minipage}
\end{figure*}

Intuitively, in both integrity and confidentiality the checked property is whether there is a path extending between a \emph{source} node and a \emph{sink} node, which does not pass through a \emph{downgrader} node. Two such paths in the example in Figure \ref{Fi:chIFS:example} are $(2,4) \longrightarrow (4,6)$ (integrity vulnerability) and  
$(2,9) \longrightarrow (9,11)$ (confidentiality vulnerability).

\subsection{The Principle of Noninterference}

An elegant way of formalizing the requirement that the source data does not affect the sink's behavior is via the principle of noninterference. This principle --- first proposed by Goguen and Meseguer in 1982 \cite{GoguenMeseguer} --- formalizes the intuition that variation ``sensitive'' values should not cause variation of ``nonsensitive'' values.

Noninterference assumes an abstract model of the program as an input/output machine. Inputs and outputs are classified as either \emph{high} or \emph{low}.
%
In confidentiality, low is interpreted as low sensitivity. A low value is thus a value that is considered unclassified. Complementarily, a high value is considered classified (or sensitive). Integrity assigns the dual meaning to high and low. 
In integrity, low means trusted, whereas high means untrusted. Thanks to these dual interpretations, both integrity and confidentiality reduce to the requirement that there be no flow of high input values to the low part of the output.
%
These dual interpretations of high and low are summarized in Table \ref{Ta:highlow}.

\begin{table}
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
					& {\bf Integrity} & {\bf Confidentiality} \\ \hline
			{\bf High}	& Untrusted & Secret \\ \hline
			{\bf Low} & Trusted & Public \\ 		
			\hline
		\end{tabular}
	\end{center}
		\caption{\label{Ta:highlow}The meaning of high and low values with the addition of downgrading}
\end{table}

With this background, we can give an informal statement of the property of noninterference. The property holds if and only if the restriction of the output to low values is the same for any pair of inputs that agree on the low values. That is, modifying the high values does not lead to any observable change in the system's output. In this sense, there is no interference between high and low values.

Formal statement of the noninterference principle requires some notation. Denote by $M$ a memory configuration, and let $M_L$ and $M_H$ denote the projection of $M$ to its low and high parts, respectively. We specialize the equality relation analogously: $M \equiv_{L} M'$ means equality over the low values between memory configurations $M$ and $M'$, and similarly, $M \equiv_{H} M'$ means equality over the high parts. Finally, let $P$ be the program, which we assume for simplicity to be deterministic, and $(P,M) \longrightarrow^{*} M'$ a complete execution of $P$ starting at memory configuration $M$ and terminating at configuration $M'$.

The definition of noninterference is then as follows:
$$
	\begin{array}{rrrcl}
	\forall M_1,M_2. &  				  & M_1 & \equiv_L & M_2 \\
							   &  \wedge & (P,M_1) & \longrightarrow^{*} & M_1' \\ 
							   &  \wedge & (P,M_2) & \longrightarrow^{*} & M_2' \\
						   & \Longrightarrow & M_1'    & \equiv_L & M_2'  \\
	\end{array}
$$
That is, if the initial memory configurations agree on their respective low parts, then the same should hold true of the final memory configurations. A visual representation of this requirements is provided in Figure \ref{noninterfere}.

\begin{figure}
	TODO!!!
	\caption{Visual representation of the noninterference principle}
\end{figure}

Let us now revisit the example in Figure \ref{Fi:chIFS:example}. First, for the integrity vulnerability extending between {\tt getParameter} and {\tt renderToHTML}, we observe that 
the user-provided data --- belonging in the high part of the input configuration --- flows into a security-sensitive operation, and so the low part of the output configuration. Analogously, the confidential --- and thus high --- data read via the {\tt readPassword} call is released to public observers of the log, thus flowing into the low part of the output configuration.

\subsection{The Problem: Enforcement of Noninterference}

While noninterference is an elegant criterion that accounts simultaneously for both integrity and confidentiality threats, it suffers from several limitations and poses some nontrivial implementation challenges. We discuss these in turn.

\subsubsection{Strictness}

A straightforward fix for the defects exhibited by the code in Figure \ref{Fi:chIFS:example} is to downgrade the high value: Sanitize or validate it to prevent integrity attacks, or declassify it to prevent release of sensitive information. Notice, however, that the noninterference property --- in its current form --- is too strict to accomodate the notion of downgrading.

Indeed, in the practical setting of production sofrware, 
whose behavior is often guided by user input and whose output often reflects data originating from multiple users, it is extremely challenging to enforce full separation between high and low values. Doing so often comes at the cost of critical functionality.

As a natural example, an e-commerce site accepts login credentials from the user, which it utilizes to access the user's account, stored on a backend database, via security-sensitive calls that could potentiall lead to an SQL injection attack. For confidentiality, public data presented to the user, such as purchase recommendations, may be influenced by sensitive data such as the user's profile.

In response, a relaxed notion of noninterference has been proposed by XYZ, which augments the basic setting with the ability to transform high values into low values via downgrading. This enhancement permits the programmer to handle security-sensitive values as they propagate through the code. With this enhancement, we arrive from Table \ref{Ta:highlow} to Table \ref{Ta:highlowExtended}.

\begin{table}
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			& {\bf Integrity} & {\bf Confidentiality} \\ \hline
			{\bf High}	& Untrusted & Secret \\ \hline
			{\bf Low} & Trusted & Public \\ \hline		
			{\bf Downgrading} & Endorsement & Declassification \\
			\hline
		\end{tabular}
	\end{center}
	\caption{\label{Ta:highlowExtended}The meaning of high and low values with the addition of downgrading}
\end{table}

\subsubsection{Correctness of Downgrading}

The introduction of downgraders as a means to accommodate information flows that transition from high to low is necessary for practical adoption of the noninterference criterion. At the same time, however, it introduces a new challenge: How do we ascertain that the downgrading operation operates correctly?

If the downgrader is incorrect, or partial, then an attacker may still be able to exploit the system. We provide an example in Figure \ref{TateExample}. This code, taken from ...

\subsubsection{Covert Channels}

So far we have implicitly considered only direct flow of information. However, security threats occupy a broader scope. In particular, an attack could become possible via indirect mechanisms, such as exploiting time or heat measurements.

In general, we refer to mechanisms for information transfer as \emph{channels}. A \emph{channel} is a means, or mechanism, to signal information through a compute device. 
\emph{Covert channels}, or \emph{side channels}, are channels that exploit a mechanism whose primary purpose is not information transfer.

Covert channels divide into several categories, which we survey in turn.

\paragraph{Implicit flows.} The control-flow behavior of a program is a means to signal information. Branching decisions governed by high values may impact low variables, which creates a covert channel.

\begin{figure}
	\begin{lstlisting}
t = h % 2;
l = 0;
if (t == 1) 
  l = 1;
	\end{lstlisting}
\caption{\label{Fi:implicitFlow}Example of an implicit flow}
\end{figure}

To illustrate this, we refer to the code in Figure \ref{Fi:implicitFlow}. In this program, we assume that {\tt l} and {\tt h} are low and high variables, respectively. Though there is no explicit assignment of the high value pointed-to by {\tt h} to {\tt l}, this program is equivalent to the program
\begin{quote}
	{\tt l = h \% 2}
\end{quote} 
where there is direct flow from high to low.

\paragraph{Termination.} Another channel, similar to implicit flows, is termination. Whether or not a program terminates is a potential channel for release of sensitive information. *** Example ***

\paragraph{Timing.} Monitoring the execution time of a program discloses information about the values it operates on. In a seminal paper \cite{XXX},

\paragraph{Probabilistic.} Under the assumption that the attacker is able to execute the program multiple times, sensitive information is potentially leaked through stochastic program properties by changing the distribution of observable data.

\paragraph{Resource exhaustion.} Another channel is exhaustion of finite compute resources, such as memory or disk space.

\paragraph{Power.} Similarly to time measurement, the power consumed by the compute device throughout the program's execution is a potential hint as to which values are being processed.

\paragraph{Sensors.} Correlations between sensors can disclose to an attacker unintended information. A recent example is use of the gyroscope built into modern smartphones to recover the speaker's identity, and even parse speech.

\section{Language-based Information-flow Security}

Language-based mechanisms are used to address different security threats. A notable example is the Java runtime system, which features bytecode verification \cite{XXX}, sandboxing \cite{XXX} as well as stack inspection \cite{XXX}. While bytecode verification requires a limited level of intra-procedural static program analysis, sanboxing and stack inspection rely solely on the Java language.

Language-based information-flow security aims to achieve exactly this. The core idea is to explode, or refine, ordinary types, like the Java {\tt String} and {\tt Integer}, into two parts. The first is the original type. The second part is a label that statically enforces security restrictions on the value via type checking, which is referred to as the \emph{security type} of the expression.

A security type system, therefore, is a set of security rules, which specify the type assigned to a program or expression given the types assigned to subexpressions. We illustrate some of the rules in Figure \cite{SabelfeldJournalPaper}.

\noindent {\bf TODO:} Add discussion of rules...

A major advantage of the language-based approach lies in its ability to account for implicit flows. This is done by assigning a label also to the program counter. Returning to the example in Figure \ref{Fi:implicitFlow}, we assign a high label to the conditional statement, which tests a high value, and consequently also to all the program counters dominated by the condition, namely the assignment to {\tt l}. This makes the implicit flow explicit.

\section{Information-flow Security via Program Analysis}

An alternative approach to annotating the program with security types is to track flow of information throughout the code via program analysis. In this scheme, there are three categories of statements that are distinguished from the rest, as mentioned briefly above:
\begin{itemize}
	\item \emph{Source} statements are statements that read either untrusted data (integrity) or sensitive data (confidentiality).
	\item \emph{Sink} statements either perform a security-sensitive operation (integrity), or release information to untrusted observers (confidentiality).
	\item Finally, \emph{downgrader} statements either sanitize/validate untrusted data (integrity), or anonymize/declassify sensitive data (confidentiality).
\end{itemize}

These three categories of statements are organized into security rules of the form $\left( Src,Dwn,Snk \right)$. $Src$, $Dwn$ and $Snk$ denote subsets of the source, sink and downgrader statements in the program, repsectively. Rules restrict the relevant security threats. Flow of information between source $src$ and sink $snk$ constitutes a security threat only if there is a rule $r$, such that $src \in r.Src$ and $snk \in r.Snk$. By the same token, the flow is benign only if there is a downgrader from $r.Dwn$ along it.

Security enforcement via program analysis computes a fixpoint solution for the dataflow behavior of the program, where the fixpoint system is seeded by the dataflow facts arising at the source statements. These are propagated through the code.

A common way of handling downgraders is to terminate the flow at their return point. This creats a challenge however, as the status of a method as a downgrader depends on the given source and sink, but the sink is not yet known at the point where the judgment whether to terminate the flow is made. Unfortunately, the simple solution of treating downgraders as globally applicable is unsound, which mandates a more expensive solution.

Indeed, an acceptable solution is to compute the fixpoint solution on a per-rule basis. Different rules are processed separately, such that flow termination at downgrading points is conservative.

In this setting, two optimizations are made possible:
\begin{enumerate}
	\item If two rules, $r_1$ and $r_2$, agree on the set of downgraders, then they can be processed simultaneously without loss of soundness.
	\item As further optimization, even if the declared sets of downgraders for $r_1$ and $r_2$ differ, the downgraders actually used by the program, denoted $Dwn_p$, may be common to both $r_1$ and $r_2$: $Dwn_p \subseteq Dwn_{r_1} \cap Dwn_{r_2}$. In this case, too, $r_1$ and $r_2$ can be processed at the same time.
\end{enumerate}

Different techniques have been proposed to perform program security analysis. These feature different precision/performance tradeoffs. We discuss a few of these techniques, at a high level, in the following.

\subsection{Taint Analysis}

Perhaps the most popular method to track flow of information throughout the program via program analysis is \emph{taint analysis}. The idea, stated simply, is to assign a \emph{taint tag} to every value resulting from a source statement, where the tag is merely a bit indicating that the value is security relevant. The taint tags are then propagated along the program's control-flow paths.

A main reason for the wide adoption of taint analysis as a framework for program-analysis-based information-flow security is its simplicity and scalability. Practical security products are required to scale to industry-level programs, which may consist of miillions of lines of code, and so an analysis that tracks minimal information is favorable.

At the same time, the limited expressiveness of taint analysis brings about accuracy challenges.

\subsection{String Analysis}

\subsection{Staged Analysis}

### 2015-03-18 SBACSystems.tex
SBAC

\chapter{Stack-Based Access Control Systems} \label{chap:SBACSystems}

Several modern software platforms---most notably, Java, Standard Edition (SE)
\cite{book:J2SESecBook,book:LiGong}, Java, Enterprise Edition (EE) \cite{book:J2EESecBook} and
Microsoft .NET Common Language Runtime (CLR) \cite{book:DotNetSecurity}---come with an embedded
mechanism for access-control enforcement known as \emph{stack inspection}. 
This chapter describes how Stack-Based Access Control (SBAC)
works.  As a reference, it focuses on the Java SBAC system, which applies
to both Java, SE and Java, EE.  Whenever
necessary, the major differences between the Java and the CLR
SBAC systems will be highlighted.

\section{Basic Concepts}

The Java
and CLR programming models are
extensively used in different kinds of Internet applications,
especially banking, retail and electronic commerce. Due to the
distributed nature of such applications, it is essential that,
when access to a restricted resource is attempted, all code
currently on the call stack is authorized to access that resource.
In Java, the
\texttt{SecurityManager}, if active, triggers access-control
enforcement by invoking \texttt{AccessController.checkPermission}.
This method takes a \texttt{Permission} object \texttt{p} as a
parameter and performs a stack walk to verify that each caller in
the current thread of execution has been granted the access right
represented by \texttt{p}. In CLR, the stack walk is performed by
the \texttt{Demand} method, which verifies that all the code
currently on the stack has been granted the access right
represented by a given \texttt{IPermission} object.

A comparison
between the Java and CLR SBAC systems is shown in Figure
\ref{fig:JVMandCLR}.  The two directed graphs in Figure
\ref{fig:JVMandCLR} represent the executions of two program
threads on a Java and CLR platform, respectively. The vertices
in the graphs represent method invocations.  If two vertices are
connected by an edge, then the method represented by
the predecessor vertex in the edge invokes the method represented
by the successor vertex.  In both platforms, when SBAC is
enforced, a \texttt{SecurityException} is thrown if any one of the
callers on the call stack does not have the appropriate access
right.
\begin{figure}[ht]
	\centering
\includegraphics[scale=0.4]{JVMandCLR-eps-converted-to.pdf}
	\caption{The Java and CLR SBAC Systems}
	\label{fig:JVMandCLR}
\end{figure}

\section{Access Rights} \label{sec:AccessRights}

Though programmatic security is possible, in both the Java and
CLR platforms access rights are preferably granted declaratively.
A system administrator grants access rights in a security policy
database that is external to the application code. This enhances
code portability and reusability.  Typically, in SBAC systems,
access rights are by default denied to all code and subjects.
Untrusted code and subjects will only be allowed to perform basic
operations that cannot harm the system.  For a restricted
operation to succeed, all the code on the thread's stack must be
explicitly granted the right to execute that operation.

In Java, access rights are represented as objects of type
\texttt{Permission}. Each \texttt{Permission} type must be a
subclass of the \texttt{Permission} abstract class.  When a
\texttt{Permission} object is constructed, it can take zero, one,
or two \texttt{String} objects as parameters.  If present, the
first parameter is called the \emph{target} of the
\texttt{Permission} object and represents the resource being
protected, whereas the second parameter is called the
\emph{action} of the \texttt{Permission} object and represents the
mode of access. The target and action are used to better qualify
the resource being protected by the \texttt{Permission} object.
For example, the following line of code can be used to construct a
\texttt{Permission} object representing the right to access the
\texttt{log.txt} file in read and write mode:

\begin{tabbing}
	\texttt{Permission p =}\\
	\texttt{   new FilePermission("log.txt",
		"read,write");}
\end{tabbing}

Given a \texttt{Permission} object \texttt{p}, \texttt{p}'s
fully-qualified \texttt{Permission} class name along with
\texttt{p}'s target and action, if any, uniquely identify the
authorization requirement represented by \texttt{p}. Therefore,
for authorization purposes, a \texttt{Permission} object \texttt{p}
can be characterized solely based on its \textit{permission
	ID}, which consists of \textt{p}'s fully-qualified class name and
the values of the \texttt{String} instances used to instantiate
\texttt{p}.  In fact, in the Java Runtime Environment (JRE)
reference implementation, authorizations are granted to programs
and principals by just listing the corresponding permission IDs in
a flat-file policy database, called the \emph{policy file}
\cite{book:J2SESecBook}.

The \texttt{Permission} class contains an abstract method,
\texttt{implies}, which itself takes a \texttt{Permission} object
as a parameter and returns a \texttt{boolean} value, and that
every non-abstract \texttt{Permission} subclass must implement. If
\texttt{p} and \texttt{q} are two \texttt{Permission} objects and
\texttt{p.implies(q)} returns \texttt{true}, then this means that
granting \texttt{p} is implicitly equivalent to having granted
\texttt{q} as well.  For example, \texttt{p} could be the
\texttt{Permission} object constructed above and \texttt{q} could
be the \texttt{Permission} object constructed with the following line of
code:

\begin{tabbing}
	\texttt{Permission q = new FilePermission("log.txt", "read");}
\end{tabbing}

If \texttt{p} is an instance of the \texttt{AllPermission} class,
then \texttt{p.implies(q)} returns \texttt{true} for any
\texttt{Permission} object \texttt{q}.

Access rights may be granted to code based on the \emph{code
	source}, which is a combination of the \emph{code base}---the
network location from which the code is coming---and the digital
certificates of the entities that digitally signed the code
\cite{book:J2SESecBook}. In Java, access rights are granted to
classes.  Each class is loaded at run time by a class loader.
Class loaders are organized as a tree, called the
\emph{class-loading delegation tree} \cite{book:J2EESecBook}. When
it loads a class, a class loader constructs a \emph{protection
	domain} characterizing the origin of the class being loaded, and
associates it with the class itself. A protection domain, which is
represented as an object of type \texttt{ProtectionDomain},
encapsulates the class' code source (represented as a
\texttt{CodeSource} object) and a \texttt{PermissionCollection}
object containing all the \texttt{Permission}s corresponding to
the access rights granted to that code source.  When it is invoked
with a \texttt{Permission} \texttt{p} as a parameter,
\texttt{checkPermission} performs a stack walk and verifies that
each of the \texttt{ProtectionDomain}s in the current thread of
execution contains in its \texttt{PermissionCollection} at least one \texttt{Permission} that implies
\texttt{p}. The set of \texttt{Permission}s effectively granted to
a thread of execution is the intersection of the sets of
\texttt{Permission}s implied by all the \texttt{ProtectionDomain}s
associated with the stack. Access-control enforcement on a CLR
platform is very similar.

From a security point of view, it should be noted that violations
of the Principle of Least Privilege may arise if unnecessary
rights are granted to code.  Conversely, if the rights granted to
code are insufficient to meet the code's security requirements,
run-time authorization failures may occur, which can cause the
application to crash.

\section{Privileged Code} \label{sec:PrivilegedCode}

Often, trusted library code has been programmed to perform
access-restricted operations---such as writing to a log file or
reading from a configuration file---that its untrusted client did
not explicitly request. Since the untrusted client will be on the
call stack when access control is enforced, the operation will not
succeed unless the client code is authorized as well.  To avoid
authorizing the client, which would constitute a violation of the
Principle of Least Privilege \cite{paper:PrincLeastPriv}, the
portion of library code performing the restricted operation can be
made \emph{privileged}.

In Java, a block of library code is made privileged by wrapping
it into a call to \texttt{AccessController.doPrivileged}. This
method takes either a \texttt{PrivilegedAction} or a
\texttt{PrivilegedExceptionAction} parameter, whose \texttt{run}
method contains the portion of code that needs to be made
privileged.  When access control is enforced, the presence of
privileged code on the call stack causes the stack walk to stop at
the stack frame corresponding to the library method calling
\texttt{doPrivileged}. As a result, the library code calling
\texttt{doPrivileged} must have been granted the right to perform
the restricted operation, but the client code is implicitly
granted that right while the current thread is executing.  A
potential problem generated by \texttt{doPrivileged} is that, if a
call to \texttt{doPrivileged} leads to multiple
\texttt{checkPermission} calls with different \texttt{Permission}s
being checked, all those \texttt{Permission}s will be implicitly
granted to client code indiscriminately.

In CLR, the mechanism for privileged code is more granular.  To
make code privileged, a trusted library code must call the
\texttt{Assert} method with an \texttt{IPermission} object
\texttt{p} as a parameter. The difference from Java's
\texttt{doPrivileged} is that \texttt{Assert} does not cause the
stack walk to stop, but instead adds \texttt{p} to the stack
frames above the library method that makes the call to
\texttt{Assert}.  Therefore, unlike \texttt{doPrivileged},
\texttt{Assert} allows specifying exactly which access right
should be implicitly granted to client code, regardless of the
\texttt{IPermission} being checked by the \texttt{Demand} method
currently on the stack.  Figure \ref{fig:privilegedcode} shows and
compares how privileged code works in the Java and CLR
platforms, respectively.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.4]{privilegedcode-eps-converted-to.pdf}
	\caption{Privileged Code in the Java and CLR SBAC Systems}
	\label{fig:privilegedcode}
\end{figure}

The Java platform offers another form of \texttt{doPrivileged}
that takes an additional \texttt{AccessControlContext} instance as
an argument.  This instance encapsulates an array of
\texttt{ProtectionDomain} objects, which are typically those
associated with the thread stack at the moment in which the
\texttt{AccessControlContext} was instantiated. When it encounters
a call to this form of \texttt{doPrivileged},
\texttt{checkPermission} stops its stack walk at the stack frame
corresponding to the method that made that \texttt{doPrivileged}
call. However, in addition, \texttt{checkPermission} performs an
authorization test on each of the \texttt{ProtectionDomain}s
encapsulated in the \texttt{AccessControlContext}.

The remainder of this section further describes the problems that
can be introduced with privileged code.

\subsection{Library Code Access Control} \label{sec:LibraryCodeAC}
When a client application makes a call into a trusted library, the
library often accesses restricted resources that the client never
intended to, nor does it need to, directly access. For instance,
assume that a Java program is authorized to open a network socket.
To do so, it invokes \texttt{createSocket} on the
\texttt{LibraryCode} library class in Figure
\ref{fig:LibraryCode1}.
\begin{figure*}[h]
	\begin{tabbing}
		\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\kill
		\texttt{import java.io.*;}\\
		\texttt{import java.net.*;}\\
		\verb"public class LibraryCode {"\\
			\>\texttt{private static String logFileName = "audit.txt";}\\
			\>\verb"public static Socket createSocket(String host, int port)"\\
			\>\>\>\verb"throws UnknownHostException, IOException {"\\
				\>\>\texttt{// Create the Socket}\\
				\>\>\texttt{Socket socket = new Socket(host, port);}\\
				\>\>\texttt{// Log the Socket operation to a file}\\
				\>\>\texttt{FileOutputStream fos = new FileOutputStream(logFileName);}\\
				\>\>\texttt{BufferedOutputStream bos = new BufferedOutputStream(fos);}\\
				\>\>\texttt{PrintStream ps = new PrintStream(bos, true);}\\
				\>\>\texttt{ps.print("Socket " + host + ":" + port);}\\
				\>\>\texttt{return socket;}\\
				\>\verb"}"\\
			\verb"}"
	\end{tabbing}
	\caption{Library Code Propagating Authorization Requirements to
		Its Clients} \label{fig:LibraryCode1}
\end{figure*}

On opening a socket on behalf of a client program, the library is
programmed to log the socket operation to a file for auditing
purposes. As shown in Figure \ref{fig:NoDoPriv}, according to the
Java SBAC model, both the library and its client will need to be
granted the \verb"FilePermission" to modify the log file and the
\verb"SocketPermission" to create the socket connection, even
though the client did not explicitly request to write to the log
file. Granting that \texttt{FilePermission} to the client code
would violate the Principle of Least Privilege.
\begin{figure*}[h]
	\centering
	\includegraphics[scale=0.4]{pistoia-figure1-eps-converted-to.pdf}
	\caption{Propagating Unnecessary Authorization Requirements to Client Code}
	\label{fig:NoDoPriv}
\end{figure*}

One way to circumvent this problem is to mark the portion of the
library code responsible for logging as privileged. On a Java
platform, this prevents the stack inspection for the log operation
from going beyond the \texttt{createSocket} method, and
temporarily exempts the client from the \texttt{FilePermission}
requirement during the execution of \texttt{createSocket}. From a
practical point of view, a Java developer must implement either
the \verb"PrivilegedExceptionAction" or \verb"PrivilegedAction"
interface, depending on whether the privileged code could throw a
checked \verb"Exception" or not, respectively.  Both these
interfaces have a \verb"run" method that, once implemented, must
contain the portion of library code performing the restricted
operation not directly requested by the client. Next, the
\verb"PrivilegedExceptionAction" or \verb"PrivilegedAction"
instance is passed as a parameter to the \verb"doPrivileged"
method, which will invoke the instance's \texttt{run} method.

\begin{figure*}
	\begin{tabbing}
		\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\kill
		\verb"import java.io.*;"\\
		\verb"import java.net.*;"\\
		\verb"import java.security.*;"\\
		\verb"public class LibraryCode2 {"\\
			\>\texttt{private static final String logFileName = "audit.txt";}\\
			\>\texttt{public static Socket createSocket(String host, int port) throws}\\
			\>\>\>\verb"UnknownHostException, IOException, PrivilegedActionException {"\\
				\>\>\texttt{// Create the Socket}\\
				\>\>\texttt{Socket socket = new Socket(host, port);}\\
				\>\>\texttt{// Log the Socket operation to a file using doPrivileged}\\
				\>\>\texttt{File f = new File(logFileName);}\\
				\>\>\texttt{PrivWriteOp op = new PrivWriteOp(host, port, f);}\\
				\>\>\texttt{FileOutputStream fos = (FileOutputStream)}\\
				\>\>\>\>\texttt{AccessController.doPrivileged(op);}\\
				\>\>\texttt{BufferedOutputStream bos = new
					BufferedOutputStream(fos);}\\
				\>\>\texttt{PrintStream ps = new PrintStream(bos, true);}\\
				\>\>\texttt{ps.print("Socket " + host + ":" + port);}\\
				\>\>\texttt{return socket;}\\
				\>\verb"}"\\
			\verb"}"\\
		\verb"class PrivWriteOp implements PrivilegedExceptionAction {"\\
			\>\texttt{private File f;}\\
			\>\verb"PrivWriteOp (File f) {"\\
				\>\>\verb"this.f = f;"\\
				\>\verb"}"\\
			\>\verb"public Object run() throws IOException {"\\
				\>\>\texttt{return new FileOutputStream(f);}\\
				\>\verb"}"\\
			\verb"}"
	\end{tabbing}
	\caption{Library Using Privileged Code} \label{fig:LibraryCode2}
\end{figure*}

Class \texttt{LibraryCode2} in Figure \ref{fig:LibraryCode2} is
obtained by modifying class \verb"LibraryCode" in Figure
\ref{fig:LibraryCode1}. The main modification consists of wrapping
the call to the \texttt{FileOutputStream} constructor in a
privileged block to prevent client code from requiring a
\verb"FilePermission".

\begin{figure*}[h]
	\centering
\includegraphics[scale=0.4]{pistoia-figure2-eps-converted-to.pdf}
	\caption{Preventing Unnecessary Authorization Requirements for Client Code}
	\label{fig:withDoPriv}
\end{figure*}

As shown in Figure \ref{fig:withDoPriv}, when the
\verb"checkPermission" method performs a stack walk to verify that
all the code in the active calling sequence has been granted the
right to write to the log file, the \verb"LibraryCode2" class will
be required to have the necessary \verb"FilePermission", but the
client code running on top of it will not.  However, the client
code will still be required to have the \verb"SocketPermission"
necessary to open the socket since the call to the \verb"Socket"
constructor in \verb"LibraryCode" is not privileged.

\subsection{Correct Placement of Privileged Code}
Taking preexisting library code and understanding which portions
of it should be made privileged by performing manual code
inspection is a difficult task, which is even more challenging
when the library is large or complex. Besides identifying the
blocks of library code that require authorizations, developers
must understand which authorizations are implicitly granted to
possibly unknown client code when the library code is made
privileged.

Frequently, libraries are not written with security as a concern,
or they are written to run on a version of the Java Runtime
Environment (JRE) prior to 1.2.\footnote{The Java fine-grained
	access control model was introduced in version 1.2.} When a Java
\texttt{SecurityManager} is finally turned on for a particular
application, \texttt{SecurityException}s are thrown due to access
control violations. It may be very difficult to understand which
portions of library code should be made privileged in order to
prevent client code from requiring unnecessary authorizations that
are only needed by the library code.

In practice, this problem is solved empirically. The developer
tests the library code with sample client code that makes calls
into the library. Typically, the client code is granted only a
limited number of authorizations, while the library is granted
sufficient authorizations, such as \verb"AllPermission". Next, it
is necessary to take note of all the \verb"SecurityException"s
generated when running the test cases, and to distinguish between
two categories of \verb"SecurityException"s:
\begin{enumerate}
	\item The \verb"SecurityException"s due to the client code's attempting to
	access some restricted resources through the library without
	the adequate authorizations
	\label{list:SecurityExceptionsOfType1}
	\item The \verb"SecurityException"s due to the library's attempting
	to access some restricted resources on its own without using
	privileged code \label{list:SecurityExceptionsOfType2}
\end{enumerate}
Eliminating a \verb"SecurityException" of category
\ref{list:SecurityExceptionsOfType2} requires inspecting the
library source code, identifying which portion of it is
responsible for accessing the protected resource, and making that
portion of code privileged. A \verb"SecurityException" of category
\ref{list:SecurityExceptionsOfType1} can instead be eliminated by
granting the client code the necessary access rights, but this
operation must be performed cautiously because granting
authorizations to the client could hide \verb"SecurityException"s
of category \ref{list:SecurityExceptionsOfType2}. Manually
performing this task is difficult, tedious, and error-prone. After
modifying the library code or the client security policy, the
developer must rerun the test cases. This process must be
repeated, possibly many times, until there are no more
authorization failures. Additionally, \verb"doPrivileged"
requirements in the library code may remain undiscovered during
testing due to an insufficient number of test cases, which makes
production code potentially unstable.

It is also important to avoid introducing ``unnecessary'' and
``redundant'' privileged blocks code.  Privileged code is
\emph{unnecessary} if there is no path from it to any security
check, and it is \emph{redundant} if all the security checks it
leads to are covered by other privileged code. Unnecessary or
redundant privileged code may lead to violations of the Principle
of Least Privilege, especially during subsequent code maintenance.
Additionally, from a performance point of view, it can be
expensive.  Therefore, such code should be made unprivileged.

\subsection{Tainted Variables}

A variable in a program is said to be \emph{tainted} if its value
can be arbitrarily set by the client code
\cite{Web:SecCodeGuidelines}. For example, if the privileged code
is responsible for writing to a log file, the name of the log file
should not be a tainted variable, or an untrusted caller could
invoke that privileged code and modify any file in the file
system.

A tainted variable is not necessarily a security problem.  In Java
2, a tainted variable may constitute a security problem if it is
also a \emph{privileged} variable, meaning that it is used inside
privileged code \cite{Web:SecCodeGuidelines}.  Even a tainted
privileged variable is not necessarily a security problem.  In
fact, it is possible to distinguish two types of privileged
tainted variables: if a privileged tainted variable is used to
access a restricted resource, it is said to be \emph{malicious},
otherwise it is said to be \emph{benign}. Since security checks
are not performed beyond the stack frame invoking
\texttt{doPrivileged}, an untrusted client application could
exploit a malicious variable to have the privileged code access
restricted resources on its behalf. Consider, for example, the
\texttt{GetSocket} utility class shown in Figure
\ref{fig:helperMethod}. Both \texttt{host} and \texttt{port} are
tainted variables, since an untrusted client can arbitrarily set
them. The fact that they are used inside privileged code to open a
socket makes them malicious and constitutes a potential security
risk.  An untrusted client, with no \texttt{SocketPermission}, can
invoke \texttt{getSocket} on the trusted library and have the
library open an arbitrary socket connection on its behalf.
Conversely, variable \texttt{userName}, though tainted and
privileged, is benign since its value is not used to access a
restricted resource.  In Figure \ref{fig:LibraryCode2}, variable
\texttt{logFileName} is not tainted because its value cannot be
set by a client application.

\begin{figure*}
	\begin{tabbing}
		\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\=\hspace*{0.25in}\kill
		\texttt{import java.net.*;}\\
		\texttt{import java.security.*;}\\
		\verb"public class GetSocket {"\\
			\>\verb"public static Socket getSocket(final String host, final int port,"\\
			\>\>\>\verb"final String userName) throws Exception {"\\
				\>\>\texttt{Socket s;}\\
				\>\>\texttt{PrivOp op = new PrivOp(host, port, userName);}\\
				\>\>\verb"try {"\\
					\>\>\>\texttt{s = (Socket) AccessController.doPrivileged(op);}\\
					\>\>\verb"}"\\
				\>\>\verb"catch (PrivilegedActionException e) {"\\
					\>\>\>\verb"throw e.getException();"\\
					\>\>\verb"}"\\
				\>\>\texttt{return s;}\\
				\>\verb"}"\\
			\verb"}"\\
		\verb"class PrivOp implements PrivilegedExceptionAction {"\\
			\>\texttt{private String host, userName;}\\
			\>\texttt{int port;}\\
			\>\verb"PrivOp(String host, int port, String userName) {"\\
				\>\>\verb"this.host = host;"\\
				\>\>\verb"this.port = port;"\\
				\>\>\verb"this.userName = userName;"\\
				\>\verb"}"\\
			\>\verb"public Object run() throws Exception {"\\
				\>\>\texttt{System.out.println("Received request from user " + userName);}\\
				\>\>\verb"return new Socket(host, port);"\\
				\>\verb"}"\\
			\verb"}"
	\end{tabbing}
	\caption{Helper \texttt{getSocket} Method with Tainted Parameters}
	\label{fig:helperMethod}
\end{figure*}

Sometimes, it may be necessary to use tainted variables inside
privileged code to access restricted resources. In such cases it
is important to perform \emph{sanity checks} on those variables to
verify that they satisfy certain preconditions
\cite{paper:AshcraftEngler2002}. For example, in the code of
Figure \ref{fig:helperMethod}, the programmer could
\emph{sanitize} \texttt{host} and \texttt{port} and make them
untainted by refusing to execute the privileged code if, for
example, the \texttt{host} value does not end with \texttt{.edu}
and the \texttt{port} value is different from \texttt{443}.

In general, manually ensuring that no tainted variables are used
inside privileged code to access restricted resources is time
consuming and error prone. It requires:
\begin{enumerate}
	\item Identifying all the unsanitized, malicious tainted
	variables (such as \texttt{host} and \texttt{port} in Figure \ref{fig:helperMethod})
	and separating them from the benign ones (such as \texttt{userName})
	\item Determining all the control- and data-flow paths in the execution of the library that
	would allow an unsanitized, malicious tainted variable to be
	used inside some privileged code to access restricted
	resources
\end{enumerate}

Determining if the code candidate to become privileged uses
unsanitized, malicious tainted variables to access restricted
resources helps when deciding whether making that code privileged
is appropriate or not. For example, the code of Figure
\ref{fig:LibraryCode1} has two instructions that could be made
privileged:
\begin{enumerate}
	\item \texttt{Socket socket = new Socket(host, port);}
	\label{list:socket}
	\item \texttt{FileOutputStream fos = new FileOutputStream(logFileName);}
	\label{list:file}
\end{enumerate}
If Instruction \ref{list:socket} is made privileged, then
parameters \verb"host" and \verb"port", which are tainted, will
constitute a security exposure since they are used to access a
restricted resource. This is an indication that Instruction
\ref{list:socket} should not be made privileged. Conversely,
parameter \verb"logFileName" is not tainted. This is an indication
that Instruction \ref{list:file} could be made privileged.

\section{Multi-Threaded Programs} \label{sec:multi-threadedprograms}
In the Java SBAC system, when
access to a restricted resource is attempted from within a child
thread, it is not just the code on the stack of the child thread
that must be granted the right to perform the access-restricted
operation, but all the code in the child thread and in all its
ancestor threads. This is done to prevent a program from
performing security-sensitive operations by simply spawning a new
thread, which would have fewer classes associated with the
run-time stack, and be therefore potentially more trusted than its
ancestor threads.

\section{Subjects-Based Authorization} \label{sec:SubjectBasedAuthorization}
Both Java and CLR allow basing authorization decisions on the user
or service, called the \emph{subject}, executing the code.  Java
Authentication and Authorization Service (JAAS) extends the Java
code-based access control framework to support subject-based
authentication and authorization
\cite{book:J2EESecBook,paper:jaas}.  In Java, a subject is
represented as an object of type \texttt{Subject}. When a subject
is authenticated, the corresponding \texttt{Subject} instance is
populated with associated identities called \emph{principals},
represented as objects of type \texttt{Principal}. A subject may
have multiple associated principals.  For example, if the subject
is a person, two of the subject's associated principals could be
the person's name and social security number.

In Java, access rights are granted to code and principals, but
not directly to subjects. The set of access rights granted to a
subject is obtained as the union of the sets of access rights
granted to the subject's authenticated principals. The JAAS
framework offers the static methods \texttt{Subject.doAs} and
\texttt{Subject.doAsPrivileged} to perform a restricted operation
with the access rights granted to a subject.

\begin{itemize}
	\item The \texttt{doAs} method takes two parameters: the first
	is a \texttt{Subject} object, and the second is either a \texttt{PrivilegedAction} or
	\texttt{PrivilegedExceptionAction} object.  The code in the
	\texttt{run} method of the \texttt{PrivilegedAction} or
	\texttt{PrivilegedExceptionAction} instance is executed with the
	intersection of the sets of \texttt{Permission}s granted to the
	code on the call stack. However, \texttt{doAs} adds the
	\texttt{Permiossion}s granted to the subject's principals to the
	stack frames subsequent to the call to \texttt{doAs}.
	\item The \texttt{doAsPrivileged} method is similar to
	\texttt{doAs}, but it takes a third parameter, an
	\texttt{AccessControlContext} object, which encapsulates an array
	of \texttt{ProtectionDomain}s. Just like \texttt{doAs}, the \texttt{doAsPrivileged}
	method also adds the \texttt{Permission}s granted to the subject's
	principals to the subsequent stack frames. However, unlike in the \texttt{doAs} case, when
	\texttt{checkPermission} is called with a \texttt{Permission}
	parameter \texttt{p} after a call to \texttt{doAsPrivileged},
	the \texttt{doAsPrivileged} predecessors are not required to
	have been granted a \texttt{Permission} that implies \texttt{p}.
	Rather, \texttt{doAsPrivileged} interrogates the
	\texttt{ProtectionDomain}s in the \texttt{AccessControlContext}
	passed to it as a parameter \cite{book:J2EESecBook}.
\end{itemize}

Violations of the Principle of Least Privilege may arise if any of
the access rights granted to a subject's principals are
unnecessary.  Conversely, if the rights granted are insufficient
to meet the security requirements of an application, run-time
authorization failures may occur, which could cause the
application to crash.

Similar to privileged code, code executed under the authority of a
subject can be ``unnecessary'' or ``redundant.'' Specifically, a
call to \texttt{doAs} or \texttt{doAsPrivileged} is
\emph{unnecessary} if it does not lead to any
\texttt{checkPermission} call, and it is \emph{redundant} if all
the \texttt{checkPermission} calls it leads to are covered by
calls to \texttt{doPrivileged}. Code unnecessarily or redundantly
executed under the authority of a subject may lead to violations
of the Principle of Least Privilege, especially during subsequent
code maintenance. Additionally, from a performance point of view,
it can be expensive.

### 2015-03-26 SBACSystems.tex
SBAC + Biblio

@@
 can be characterized solely based on its \textit{permission
-	ID}, which consists of \textt{p}'s fully-qualified class name and
+	ID}, which consists of \texttt{p}'s fully-qualified class name and
 the values of the \texttt{String} instances used to instantiate

### 2015-03-26 infoflow.tex
distribution

@@
 
-An elegant way of formalizing the requirement that the source data does not affect the sink's behavior is via the principle of noninterference.
-This principle --- first proposed by Goguen and Meseguer in 1982 \cite{GoguenMeseguer} --- formalizes the intuition that variation ``sensitive'' values should not cause variation of ``nonsensitive'' values.
-
-Noninterference assumes an abstract model of the program as an input/output machine. Inputs and outputs are classified as either \emph{high} or \emph{low}.
-%
-In confidentiality, low is interpreted as low sensitivity.
-A low value is thus a value that is considered unclassified.
-Complementarily, a high value is considered classified (or sensitive).
-Integrity assigns the dual meaning to high and low.
-In integrity, low means trusted, whereas high means untrusted.
-Thanks to these dual interpretations, both integrity and confidentiality reduce to the requirement that there be no flow of high input values to the low part of the output.
-%
-These dual interpretations of high and low are summarized in Table \ref{Ta:highlow}.
-
-\begin{table}
-	\begin{center}
-		\begin{tabular}{|c|c|c|}
-			\hline
-					& {\bf Integrity} & {\bf Confidentiality} \\ \hline
-			{\bf High}	& Untrusted & Secret \\ \hline
-			{\bf Low} & Trusted & Public \\ 		
-			\hline
-		\end{tabular}
-	\end{center}
-		\caption{\label{Ta:highlow}The meaning of high and low values with the addition of downgrading}
-\end{table}
-
-With this background, we can give an informal statement of the property of noninterference.
-The property holds if and only if the restriction of the output to low values is the same for any pair of inputs that agree on the low values.
-That is, modifying the high values does not lead to any observable change in the system's output.
-In this sense, there is no interference between high and low values.
-
-Formal statement of the noninterference principle requires some notation.
-Denote by $M$ a memory configuration, and let $M_L$ and $M_H$ denote the projection of $M$ to its low and high parts, respectively.
-We specialize the equality relation analogously: $M \equiv_{L} M'$ means equality over the low values between memory configurations $M$ and $M'$, and similarly, $M \equiv_{H} M'$ means equality over the high parts.
-Finally, let $P$ be the program, which we assume for simplicity to be deterministic, and $(P,M) \longrightarrow^{*} M'$ a complete execution of $P$ starting at memory configuration $M$ and terminating at configuration $M'$.
-
-The definition of noninterference is then as follows:
-$$
-	\begin{array}{rrrcl}
-	\forall M_1,M_2. &  				  & M_1 & \equiv_L & M_2 \\
-							   &  \wedge & (P,M_1) & \longrightarrow^{*} & M_1' \\ 
-							   &  \wedge & (P,M_2) & \longrightarrow^{*} & M_2' \\
-						   & \Longrightarrow & M_1'    & \equiv_L & M_2'  \\
-	\end{array}
-$$
-That is, if the initial memory configurations agree on their respective low parts, then the same should hold true of the final memory configurations.
-A visual representation of this requirements is provided in Figure \ref{noninterfere}.
-
-\begin{figure}
-	TODO!!!
-	\caption{Visual representation of the noninterference principle}
-\end{figure}
-
-Let us now revisit the example in Figure \ref{Fi:chIFS:example}. First, for the integrity vulnerability extending between {\tt getParameter} and {\tt renderToHTML}, we observe that 
-the user-provided data --- belonging in the high part of the input configuration --- flows into a security-sensitive operation, and so the low part of the output configuration.
-Analogously, the confidential --- and thus high --- data read via the {\tt readPassword} call is released to public observers of the log, thus flowing into the low part of the output configuration.
-
 \subsection{The Problem: Enforcement of Noninterference}
-
-While noninterference is an elegant criterion that accounts simultaneously for both integrity and confidentiality threats, it suffers from several limitations and poses some nontrivial implementation challenges.
-We discuss these in turn.
-
-\subsubsection{Strictness}
-
-A straightforward fix for the defects exhibited by the code in Figure \ref{Fi:chIFS:example} is to downgrade the high value: Sanitize or validate it to prevent integrity attacks, or declassify it to prevent release of sensitive information.
-Notice, however, that the noninterference property --- in its current form --- is too strict to accomodate the notion of downgrading.
-
-Indeed, in the practical setting of production sofrware, 
-whose behavior is often guided by user input and whose output often reflects data originating from multiple users, it is extremely challenging to enforce full separation between high and low values.
-Doing so often comes at the cost of critical functionality.
-
-As a natural example, an e-commerce site accepts login credentials from the user, which it utilizes to access the user's account, stored on a backend database, via security-sensitive calls that could potentiall lead to an SQL injection attack.
-For confidentiality, public data presented to the user, such as purchase recommendations, may be influenced by sensitive data such as the user's profile.
-
-In response, a relaxed notion of noninterference has been proposed by XYZ, which augments the basic setting with the ability to transform high values into low values via downgrading.
-This enhancement permits the programmer to handle security-sensitive values as they propagate through the code.
-With this enhancement, we arrive from Table \ref{Ta:highlow} to Table \ref{Ta:highlowExtended}.
-
-\begin{table}
-	\begin{center}
-		\begin{tabular}{|c|c|c|}
-			\hline
-			& {\bf Integrity} & {\bf Confidentiality} \\ \hline
-			{\bf High}	& Untrusted & Secret \\ \hline
-			{\bf Low} & Trusted & Public \\ \hline		
-			{\bf Downgrading} & Endorsement & Declassification \\
-			\hline
-		\end{tabular}
-	\end{center}
-	\caption{\label{Ta:highlowExtended}The meaning of high and low values with the addition of downgrading}
-\end{table}
-
-\subsubsection{Correctness of Downgrading}
-
-The introduction of downgraders as a means to accommodate information flows that transition from high to low is necessary for practical adoption of the noninterference criterion.
-At the same time, however, it introduces a new challenge: How do we ascertain that the downgrading operation operates correctly?
-
-If the downgrader is incorrect, or partial, then an attacker may still be able to exploit the system. We provide an example in Figure \ref{TateExample}. This code, taken from ...
-
-\subsubsection{Covert Channels}
-
-So far we have implicitly considered only direct flow of information.
-However, security threats occupy a broader scope.
-In particular, an attack could become possible via indirect mechanisms, such as exploiting time or heat measurements.
-
-In general, we refer to mechanisms for information transfer as \emph{channels}. A \emph{channel} is a means, or mechanism, to signal information through a compute device. 
-\emph{Covert channels}, or \emph{side channels}, are channels that exploit a mechanism whose primary purpose is not information transfer.
-
-Covert channels divide into several categories, which we survey in turn.
-
-\paragraph{Implicit flows.} The control-flow behavior of a program is a means to signal information.
-Branching decisions governed by high values may impact low variables, which creates a covert channel.
-
-\begin{figure}
-	\begin{lstlisting}
-t = h % 2;
-l = 0;
-if (t == 1) 
-  l = 1;
-	\end{lstlisting}
-\caption{\label{Fi:implicitFlow}Example of an implicit flow}
-\end{figure}
-
-To illustrate this, we refer to the code in Figure \ref{Fi:implicitFlow}.
-In this program, we assume that {\tt l} and {\tt h} are low and high variables, respectively.
-Though there is no explicit assignment of the high value pointed-to by {\tt h} to {\tt l}, this program is equivalent to the program
-\begin{quote}
-	{\tt l = h \% 2}
-\end{quote} 
-where there is direct flow from high to low.
-
-\paragraph{Termination.} Another channel, similar to implicit flows, is termination. Whether or not a program terminates is a potential channel for release of sensitive information. *** Example ***
-
-\paragraph{Timing.} Monitoring the execution time of a program discloses information about the values it operates on. In a seminal paper \cite{XXX},
-
-\paragraph{Probabilistic.} Under the assumption that the attacker is able to execute the program multiple times, sensitive information is potentially leaked through stochastic program properties by changing the distribution of observable data.
-
-\paragraph{Resource exhaustion.} Another channel is exhaustion of finite compute resources, such as memory or disk space.
-
-\paragraph{Power.} Similarly to time measurement, the power consumed by the compute device throughout the program's execution is a potential hint as to which values are being processed.
-
-\paragraph{Sensors.} Correlations between sensors can disclose to an attacker unintended information.
-A recent example is use of the gyroscope built into modern smartphones to recover the speaker's identity, and even parse speech.
 
@@
 This makes the implicit flow explicit.
-
-\section{Information-flow Security via Program Analysis}
-
-An alternative approach to annotating the program with security types is to track flow of information throughout the code via program analysis.
-In this scheme, there are three categories of statements that are distinguished from the rest, as mentioned briefly above:
-\begin{itemize}
-	\item \emph{Source} statements are statements that read either untrusted data (integrity) or sensitive data (confidentiality).
-	\item \emph{Sink} statements either perform a security-sensitive operation (integrity), or release information to untrusted observers (confidentiality).
-	\item Finally, \emph{downgrader} statements either sanitize/validate untrusted data (integrity), or anonymize/declassify sensitive data (confidentiality).
-\end{itemize}
-
-These three categories of statements are organized into security rules of the form $\left( Src,Dwn,Snk \right)$.
-$Src$, $Dwn$ and $Snk$ denote subsets of the source, sink and downgrader statements in the program, repsectively.
-Rules restrict the relevant security threats.
-Flow of information between source $src$ and sink $snk$ constitutes a security threat only if there is a rule $r$, such that $src \in r.Src$ and $snk \in r.Snk$.
-By the same token, the flow is benign only if there is a downgrader from $r.Dwn$ along it.
-
-Security enforcement via program analysis computes a fixpoint solution for the dataflow behavior of the program, where the fixpoint system is seeded by the dataflow facts arising at the source statements.
-These are propagated through the code.
-
-A common way of handling downgraders is to terminate the flow at their return point.
-This creats a challenge however, as the status of a method as a downgrader depends on the given source and sink, but the sink is not yet known at the point where the judgment whether to terminate the flow is made.
-Unfortunately, the simple solution of treating downgraders as globally applicable is unsound, which mandates a more expensive solution.
-
-Indeed, an acceptable solution is to compute the fixpoint solution on a per-rule basis. Different rules are processed separately, such that flow termination at downgrading points is conservative.
-
-In this setting, two optimizations are made possible:
-\begin{enumerate}
-	\item If two rules, $r_1$ and $r_2$, agree on the set of downgraders, then they can be processed simultaneously without loss of soundness.
-	\item As further optimization, even if the declared sets of downgraders for $r_1$ and $r_2$ differ, the downgraders actually used by the program, denoted $Dwn_p$, may be common to both $r_1$ and $r_2$: $Dwn_p \subseteq Dwn_{r_1} \cap Dwn_{r_2}$.
-In this case, too, $r_1$ and $r_2$ can be processed at the same time.
-\end{enumerate}
-
-Different techniques have been proposed to perform program security analysis.
-These feature different precision/performance tradeoffs.
-We discuss a few of these techniques, at a high level, in the following.
-
-\subsection{Taint Analysis}
-
-Perhaps the most popular method to track flow of information throughout the program via program analysis is \emph{taint analysis}.
-The idea, stated simply, is to assign a \emph{taint tag} to every value resulting from a source statement, where the tag is merely a bit indicating that the value is security relevant.
-The taint tags are then propagated along the program's control-flow paths.
-
-A main reason for the wide adoption of taint analysis as a framework for program-analysis-based information-flow security is its simplicity and scalability.
-Practical security products are required to scale to industry-level programs, which may consist of miillions of lines of code, and so an analysis that tracks minimal information is favorable.
-
-At the same time, the limited expressiveness of taint analysis brings about accuracy challenges.
-
-\subsection{String Analysis}
-
-\subsection{Staged Analysis}

### 2015-04-06 p3infoflow.tex
p3

An alternative approach to annotating the program with security types is to track flow of information throughout the code via program analysis. In this scheme, there are three categories of statements that are distinguished from the rest, as mentioned briefly above:
\begin{itemize}
	\item \emph{Source} statements are statements that read either untrusted data (integrity) or sensitive data (confidentiality).
	\item \emph{Sink} statements either perform a security-sensitive operation (integrity), or release information to untrusted observers (confidentiality).
	\item Finally, \emph{downgrader} statements either sanitize/validate untrusted data (integrity), or anonymize/declassify sensitive data (confidentiality).
\end{itemize}

These three categories of statements are organized into security rules of the form $\left( Src,Dwn,Snk \right)$. $Src$, $Dwn$ and $Snk$ denote subsets of the source, sink and downgrader statements in the program, repsectively. Rules restrict the relevant security threats. Flow of information between source $src$ and sink $snk$ constitutes a security threat only if there is a rule $r$, such that $src \in r.Src$ and $snk \in r.Snk$. By the same token, the flow is benign only if there is a downgrader from $r.Dwn$ along it.

Security enforcement via program analysis computes a fixpoint solution for the dataflow behavior of the program, where the fixpoint system is seeded by the dataflow facts arising at the source statements. These are propagated through the code.

A common way of handling downgraders is to terminate the flow at their return point. This creats a challenge however, as the status of a method as a downgrader depends on the given source and sink, but the sink is not yet known at the point where the judgment whether to terminate the flow is made. Unfortunately, the simple solution of treating downgraders as globally applicable is unsound, which mandates a more expensive solution.

Indeed, an acceptable solution is to compute the fixpoint solution on a per-rule basis. Different rules are processed separately, such that flow termination at downgrading points is conservative.

In this setting, two optimizations are made possible:
\begin{enumerate}
	\item If two rules, $r_1$ and $r_2$, agree on the set of downgraders, then they can be processed simultaneously without loss of soundness.
	\item As further optimization, even if the declared sets of downgraders for $r_1$ and $r_2$ differ, the downgraders actually used by the program, denoted $Dwn_p$, may be common to both $r_1$ and $r_2$: $Dwn_p \subseteq Dwn_{r_1} \cap Dwn_{r_2}$. In this case, too, $r_1$ and $r_2$ can be processed at the same time.
\end{enumerate}

Different techniques have been proposed to perform program security analysis. These feature different precision/performance tradeoffs. We discuss a few of these techniques, at a high level, in the following.

### 2015-04-06 Sample-7x9.tex
missing

\documentclass[timesfont]{MITPress-diacriTech-7x9}
\usepackage{xypic}

\begin{document}

\frontmatter

\halftitle{Usable Program Security Analysis}

\begin{seriespage}

\seriestitle{Series Title}

\serieseditor{Series Editor}

\begin{seriesentry}
\item Series entry title
\seriesauthor{Author 1}

\item Series entry title
\seriesauthor{Author 2}

\end{seriesentry}

\end{seriespage}

\booktitle{Usable Program Security Analysis}

\subtitle{Subtitle}

\edition{First Edition}

\edauthor{Marco Pistoia\\
	Omer Tripp
}

\MITimprint{The MIT Press\\
Cambridge, Massachusetts\\
London, England}

\begin{copyrightpage}
$\copyright$ 2009 Massachusetts Institute of Technology\\

All rights reserved. No part of this book may be reproduced in any form or by any electronic or mechanical means
(including photocopying, recording, or information storage and retrieval) without permission in writing from the
publisher.\\

For information about special quantity discounts, please email special sales@mitpress.mit.edu.\\

This book was set in Times Roman and Mathtime Pro 2 by the authors.\\

Printed and bound in the United States of America.\\

Library of Congress Cataloging-in-Publication Data\\

Introduction to algorithms / Thomas H. Cormen $\ldots$ [et al.].ó3rd ed.\\
\cindent p. cm.\\

Includes bibliographical references and index.\\
ISBN 978-0-262-03384-8 (hardcover : alk. paper)óISBN 978-0-262-53305-8 (pbk. : alk. paper)\\
1. Computer programming. 2. Computer algorithms. I. Cormen, Thomas H.\\

QA76.6.I5858 2009\\
005.1ódc22\\

\hspace*{12pc}2009008593\\

10\ 9\ 8\ 7\ 6\ 5\ 4\ 3

\end{copyrightpage}

\dedication{Dedication text}

\begin{epigraphpage}
On 4 April 1980, following a conflict with the Peruvian
government, Fidel Castro ordered the guards posted in front of the Peruvian
embassy in Havana withdrawn. Seizing the chance offered by this absence,
almost 11,000 Cubans stormed into the embassy and demanded political asylum.
Images of a multitude of hungry and thirsty men, women, and children who
were even perched in trees and on the roof of the embassy, were immediately
broadcast worldwide.\source{Source}
\end{epigraphpage}

\tableofcontents

\listoffigures

\listoftables

\begin{contributors}

\name{Thomas H. Cormen}
\affil{Professor and Chair, Department of Computer Science,  Massachusetts Institute of Technology}

\name{Ronald L. Rivest}
\affil{CSAIL, 32 Vassar Street, Room 32-G692, Cambridge MA 02139}

\end{contributors}

\mainmatter

\part{Role of Algorithms}

\input{infoflow}

\chapter{The Role of Algorithms in Computing}

\author[1]{Thomas H. Cormen}

\affil[1]{Professor and Chair, Department of Computer Science,  Massachusetts Institute of Technology}

\maketitle

\epigraph{We begin by examining the insertion sort algorithm to solve the sorting problem introduced in
Chapter 1. We define a ``pseudocode'' that should be familiar to you if you have done computer
programming, and we use it to show how we shall specify our algorithms. Having specified the insertion
sort algorithm, we then argue that it correctly sorts, and we analyze its running time. The analysis
introduces a notation that focuses on how that time increases with the number of items to be sorted.
Following our discussion of insertion sort, we introduce the divide-and-conquer approach to the design of
algorithms and use it to develop an algorithm called merge sort. We end with an analysis of merge sortís
running time.}{Test sentence}

\noindent What are algorithms? Why is the study of algorithms worthwhile? What is the role of algorithms
relative to other technologies used in computers? In this chapter, we will answer these questions.

\Ahead{INTRODUCTION}

This part will start you thinking about designing and analyzing algorithms. It is intended to be a gentle
introduction to how we specify algorithms, some of the design strategies we will use throughout this book,
and many of the fundamental ideas used in algorithm analysis. Later parts of this book will build upon
this base. Chapter 1 provides an overview of algorithms and their place in modem computing systems. This
chapter defines what an algorithm is and lists some examples. It also makes a case that we should consider
algorithms as a technology, alongside technologies such as fast hardware, graphical user interfaces,
object-oriented systems, and networks.

\chapter{Getting Started}

\begin{abstract}
This chapter will familiarize you with the framework we shall use throughout the book to think about the
design and analysis of algorithms. It is self-contained, but it does include several references to
material that we introduce in Chapters 3 and 4. (It also contains several summations, which Appendix A
shows how to solve.)
\end{abstract}

\noindent We begin by examining the insertion sort algorithm to solve the sorting problem introduced in
Chapter 1. We define a ``pseudocode'' that should be familiar to you if you have done computer
programming, and we use it to show how we shall specify our algorithms. Having specified the insertion
sort algorithm, we then argue that it correctly sorts, and we analyze its running time. The analysis
introduces a notation that focuses on how that time increases with the number of items to be sorted.
Following our discussion of insertion sort, we introduce the divide-and-conquer approach to the design of
algorithms and use it to develop an algorithm called merge sort. We end with an analysis of merge sort's
running time.

\section{Insertion Sort}

The numbers that we wish to sort are also known as the \textit{keys}. Although conceptually we are sorting
a sequence, the input comes to us in the form of an array with $n$ elements.

In this book, we shall typically describe algorithms as programs written in a \textit{pseudocode} that is
similar in many respects to $\mathrm{C},\ \mathrm{C}++$, Java, Python, or Pascal. If you have been
introduced to any of these languages, you should have little trouble reading our algorithms. What
separates pseudocode from ``real'' code is that in pseudocode, we employ whatever expressive method is
most clear and concise to specify a given algorithm. Sometimes, the clearest method is English, so do not
be surprised if you come across an English phrase or sentence embedded within a section of ``real'' code.
Another difference between pseudocode and real code is that pseudocode is not typically concerned with
issues of software engineering. Issues data abstraction, modularity, and error handling are often ignored
in order to convey the essence of the algorithm more concisely.

\subsection{The Divide-and-Conquer Approach}

Many useful algorithms are \textit{recursive} in structure: to solve a given problem, they call themselves
recursively one or more times to deal with closely related subproblems. These algorithms typically follow
a \textit{divide-and-conquer} approach: they break the problem into several subproblems that are similar
to the original problem but smaller in size, solve the subproblems recursively, and then combine these
solutions to create a solution to the original problem.

\subsubsection{Order of Growth}

We used some simplifying abstractions to ease our analysis of the INSERTIONSORT procedure. First, we
ignored the actual cost of each statement, using the constants $c_{i}$ to represent these costs. Then, we
observed that even these constants give us more detail than we really need: we expressed the worst-case
running time as $an^{2}+bn+c$ for some constants $a,\ b$, and $c$ that depend on the statement costs
$c_{i}$. We thus ignored not only the actual statement costs, but also the abstract costs $c_{i}$.

We shall now make one more simplifying abstraction: it is the \textit{rate of growth}, or \textit{order of
growth}, of the running time that really interests us. We therefore consider only the leading term of a
formula $(\mathrm{e}.\mathrm{g}.,\ an^{2})$, since the lower-order terms are relatively insignificant for
large values of $n$. We also ignore the leading term's constant coefficient, since constant factors are
less significant than the rate of growth in determining computational efficiency for large inputs. For
insertion sort, when we ignore the lower-order terms and the leading term's constant coefficient, we are
left with the factor of $n^{2}$ from the leading term. We write that insertion sort has a worst-case
running time of $\Theta(n^{2})$ (pronounced ``theta of n-squared``). We shall use $\Theta$-notation
informally in this chapter, and we will define it precisely in Chapter 3.

\paragraph{Analysis of Insertion Sort}

Since the number atoms in the observable universe is estimated to be about $10^{80}$, which is much less
than $2^{65536}$, we rarely encounter an input size $n$ such that $\mathrm{lg}^{*}n>5$. Fibonacci numbers
are related to the \textit{golden ratio} $\phi$ and to its conjugate $\hat{\phi}$, which are the two roots
of the equation.

\subparagraph{Analysis of Merge Sort}

Although the pseudocode for MERGE-SORT works correctly when the number of elements is not even, our
recurrence-based analysis is simplified if we assume that the original problem size is a power of 2. Each
divide step then yields two subsequences of size exactly $n/2$. In Chapter 4, we shall see that this
assumption does not affect the order of growth of the solution to the recurrence.

\subsubparagraph*{Analyzing Divide-and-Conquer Algorithms}

We reason as follows to set up the recurrence for $T(n)$, the worst-case running time of merge sort on $n$
numbers. Merge sort on just one element takes constant time. When we have $n>1$ elements, we break down
the running time as follows.

\section{Asymptotic Notation}

The notations we use to describe the asymptotic running time of an algorithm are defined in terms of
functions whose domains are the set of natural numbers $\mathbb{N}=\{0,1,2,\ \ldots\}$. Such notations are
convenient for describing the worst-case running-time function $T(n)$, which usually is defined only on
integer input sizes. We sometimes find it convenient, however, to \textit{abuse} asymptotic notation in a
variety of ways. For example, we might extend the notation to the domain of real numbers or,
alternatively, restrict it to a subset of the natural numbers. We should make sure, however, to understand
the precise meaning of the notation so that when we abuse, we do not \textit{misuse} it. This section
defines the basic asymptotic notations and also introduces some common abuses.

\begin{theorem}\label{thm1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{theorem}

In practice, rather than using Theorem~\ref{thm1} to obtain asymptotic upper and lower bounds from
asymptotically tight bounds, as we did for this example, we usually use it to prove asymptotically tight
bounds from asymptotic upper and lower bounds in \ref{cor1}. The looping constructs while, for, and
repeat-until and the if-else conditional construct have interpretations similar to those in $\mathrm{C},\
\mathrm{C}++$, Java, Python, and Pascal. In this book, the loop counter retains its value after exiting
the loop, unlike some situations that arise in $\mathrm{C}++$, Java, and Pascal. We shall use this method
of loop invariants to show correctness later in this chapter and in other chapters as well.

\begin{corollary}\label{cor1}
Prove that the running time of an algorithm is $\Theta(g(n))$ if and only if its worst-case running time
is $O(g(n))$ and its best-case running time is $\Omega(g(n))$.
\end{corollary}

%%
When the first two properties hold, the loop invariant is true prior to every iteration of the loop. (Of
course, we are free to use established facts other than the loop invariant itself to prove that the loop
invariant remains true before each iteration.) We prefer not to get bogged down in such formalism, and so
we rely on our informal analysis to show that the second property holds for the outer loop. We shall use
this method of loop invariants to show correctness later in this chapter and in other chapters as well. We
prefer not to get bogged down in such formalism, and so we rely on our informal analysis to show that the
second property holds for the outer loop. We shall use this method of loop invariants to show correctness
later in this chapter and in other chapters as well in Definition~\ref{defi1}.

\begin{definition}\label{defi1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{definition}

Note the similarity to mathematical induction, where to prove that a property holds, you prove a base case
and an inductive step. Here, showing that the invariant holds before the first iteration corresponds to
the base case, and showing that the invariant holds from iteration to iteration corresponds to the
inductive step.

\begin{proposition}\label{prop1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{proposition}

The third property is perhaps the most important one, since we are using the loop invariant to show
correctness.\footnote{Not all authors define the asymptotic notations in the same way, although the
various definitions agree in most common situations. Some of the alternative definitions encompass
functions that are not asymptotically nonnegative, as long as their absolute values are appropriately
bounded. Other properties of elementary mathematical functions can be found in any good mathematical
reference, contain a wealth material on discrete mathematics as used in computer science.} Typically, we
use the loop invariant along with the condition that caused the loop to terminate. The termination
property differs from how we usually use mathematical induction, in which we apply the inductive step
infinitely; here, we stop the ``induction'' when the loop terminates. Let us see how these properties hold
for insertion sort. We start by showing that the loop invariant holds before the first loop iteration.

%%

\begin{lemma}\label{lem1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{lemma}

We start by showing that the loop invariant holds before the first loop iteration, when $j=2^{1}$ The
subarray $A[1.\ .\ j-1]$, therefore, consists just the single element $A[1]$, which is in fact the
original element in $A[1]$. Moreover, this subarray is sorted (trivially, of course), which shows that the
loop invariant holds prior to the first iteration of the loop.\footnote{It is unlikely that the same constant exactly represents both the time to solve problems of size 1
and the time per array element of the divide and combine steps.}

\begin{assumption}\label{assum1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{assumption}

A more formal treatment of the second property would require us to state and show a loop invariant for the
while loop of lines 5-7. At this point, however, when the loop is a for loop, the moment at which we check
the loop invariant just prior to the first iteration is immediately after the initial assignment to the
loop-counter variable and just before the first test in the loop header. In the case of INSERTIONSORT,
this time is after assigning 2 to the variable $j$ but before the first test of whether $j\leq A$.
\textit{length}.

\begin{rules}\label{rul1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{rules}

We prefer not to get bogged down in such formalism, and so we rely on our informal analysis to show that
the second property holds for the outer loop. We shall use this method of loop invariants to show
correctness later in this chapter and in other chapters as well.

\begin{examples}
Prove that the running time of an algorithm is $\Theta(g(n))$ if and only if its worst-case running time
is $O(g(n))$ and its best-case running time is $\Omega(g(n))$.
\end{examples}

Finally, we examine what happens when the loop terminates. The condition causing the for loop to terminate
is that $j>A$. \textit{length} $=n$. Because each loop iteration increases $j$ by 1, we must have $j=n+1$
at that time. Substituting $n+1$ for $j$ in the wording of loop invariant, we have that the subarray
$A[1.\ .n]$ consists of the elements originally in $A[1.\ .n]$, but in sorted order. Observing that the
subarray $A[1.\ .n]$ is the entire array, we conclude that the entire array is sorted. Hence, the
algorithm is correct.

\section{Pseudocode Conventions}

We use the following conventions in our pseudocode.

Indentation indicates block structure. For example, the body of the for loop that begins on line 1
consists of lines 2-8, and the body of the while loop that begins on line 5 contains lines 6-7 but not
line 8. Our indentation style applies to if-else statements as well. Using indentation instead of
conventional indicators of block structure, such as begin and end statements, greatly reduces clutter
while preserving, or even enhancing, clarity.

\begin{remark}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{remark}

The looping constructs while, for, and repeat-until and the if-else conditional construct have
interpretations similar to those in $\mathrm{C},\ \mathrm{C}++$, Java, Python, and Pascal. In this book,
the loop counter retains its value after exiting the loop, unlike some situations that arise in
$\mathrm{C}++$, Java, and Pascal. Thus, \hbox{immediately} after a for loop, the loop counter`s value is
the value that first exceeded the for loop bound. We used this property in our correctness argument for
insertion sort. The for loop header in line 1 is for $j=2$ to \textit{A. length}, and so when this loop
terminates, $j=A. length+1$ (or, equivalently, $j=n+1$, since $n=A$. \textit{length}). We use the keyword
to when a for loop increments its loop counter in each iteration, and we use the keyword downto when a for
loop decrements its loop counter. %When the loop counter changes by an amount greater than 1, the amount of
%change follows the optional keyword by.

\begin{demonstrations}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{demonstrations}

A multiple assignment of the form $i=j=e$ assigns to both variables $i$ and $j$ the value of expression
$e$; it should be treated as equivalent to the assignment $j=e$ followed by the assignment $i=j$.

\begin{solution}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{solution}

Variables (such as $i,\ j$, and \textit{key}) are local to the given procedure. We shall not use global
variables without explicit indication.

\begin{proof}
We use the bounds in Lemma 4.3 to evaluate the summation (4.21) from Lemma 4.2. For case 1, we have
\begin{align*}
T(n) &= \Theta(n^{\log_{b}a})+O(n^{\log_{b}a}) \\
      &= \Theta(n^{\log_{b}a}),
\end{align*}
and for case 2,
\begin{align*}
T(n) &= \Theta(n^{\log_{b}a})+\Theta(n^{\log_{b}a} \mathrm{lg}n)\\
&= \Theta(n^{\log_{b}a}\mathrm{lg}n).
\end{align*}
For case 3,
\begin{align*}
T(n) &= \Theta(n^{\log_{b}a})+\Theta(f(n))\\
     &= \Theta(f(n)),
\end{align*}
because $f(n)=\Omega(n^{\log_{b}a+\epsilon})$.
\end{proof}

We access array elements by specifying the array name followed by the index in square brackets. For
example, $A[i]$ indicates the $i$ th element of the array $A$. The notation ``$\ldots$'' is used to
indicate a range of values within an array. Thus, $A[1\ldots j]$ indicates the subarray of $A$ consisting
of the $j$ elements $A[1],\ A[2],\ \ldots,\ A[j]$.

\section{What Kinds of Problems are Solved by Algorithms?}

\subsection{Numbered Lists}

These lists are far from exhaustive (as you again have probably surmised from this book's heft), but
exhibit two characteristics that are common to many interesting algorithmic problems:
\begin{enumerate}[4.]

\item They have many candidate solutions, the overwhelming majority of which do not solve the problem at hand.
Finding one that does, or one that is ``best,'' can present quite a challenge.

\item
They have practical applications. Of the problems in the above list, finding the shortest path provides
the easiest examples. A transportation firm, such as a trucking or railroad company, has a financial
interest in finding shortest paths through a road or rail network because taking shorter paths results in
lower labor and fuel costs. Or a routing node on the Internet may need to find the shortest path through
the network in order to route a message quickly. Or a person wishing to drive from New York to Boston may
want to find driving directions from an appropriate Web site, or she may use her GPS while driving.

\begin{enumerate}[a.]

\item No black-heights in the tree have changed.

\item No red nodes have been made adjacent. Because $y$ takes $z$'s place in the tree, along with $z$'s color,
we cannot have two adjacent red nodes at $y$'s new position in the tree. In addition, if $y$ was not $z$'s
right child, then $y$'s original right child $x$ replaces $y$ in the tree. If $y$ is red, then $x$ must be
black, and so replacing $y$ by $x$ cannot cause two red nodes to become adjacent.

\item Since $y$ could not have been the root if it was red, the root remains black.

\begin{enumerate}[iii.]

\item
Otherwise, $u=2^{2^{k}}$ for some integer $k\geq 1$, so that $u\geq 4$. In addition to the universe size
$u$, the data structure \textit{pro to-vEB} $(u)$ contains the following attributes, illustrated.

\item An array \textit{cluster} $[0.\ .\ \sqrt{u}-1]$ of $\sqrt{u}$ pointers, each to a \textit{pro to-vEB}
$(\sqrt{u})$ structure.

\item If $u=2$, then it is the base size, and it contains an array $A[0.\ .\ 1]$ of two bits.
\end{enumerate}

\item Since $y$ could not have been the root if it was red, the root remains black.

\end{enumerate}

\item Select a data structure that you have seen previously, and discuss its strengths and limitations.

\item  How are the shortest-path and traveling-salesman problems given above similar? How are they different?
Come up with a real-world problem in which only the best solution will do. Then come up with one in which
a solution that is ``approximately'' the best is good enough.
\end{enumerate}

\subsection{Data Structures}

This book also contains several data structures. A \textit{data structure} is a way to store and organize
data in order to facilitate access and modifications. No single data structure works well for all
purposes, and so it is important to know the strengths and limitations of several of them.

\subsection{Unnumbered Lists}

These lists are far from exhaustive (as you again have probably surmised from this book's heft), but
exhibit two characteristics that are common to many interesting algorithmic problems:
\begin{unlist}
\item They have many candidate solutions, the overwhelming majority of which do not solve the problem at hand.
Finding one that does, or one that is ``best,'' can present quite a challenge.

\item
They have practical applications. Of the problems in the above list, finding the shortest path provides
the easiest examples. A transportation firm, such as a trucking or railroad company, has a financial
interest in finding shortest paths through a road or rail network because taking shorter paths results in
lower labor and fuel costs. Or a routing node on the Internet may need to find the shortest path through
the network in order to route a message quickly.

\end{unlist}

When an algorithm contains a recursive call to itself, we can often describe its running time by a {\it
recurrence equation} or {\it recurrence}, which describes the overall running time on a problem of size
$n$ in terms of the running time on smaller inputs. We can then use mathematical tools to solve the
recurrence and provide bounds on the performance of the algorithm.

\subsection{Bullet List}

Sorting is by no means the only computational problem for which algorithms have been developed. (You
probably suspected as much when you saw the size of this book.) Practical applications of algorithms are
ubiquitous and include the following examples:
\begin{itemize}
\item The Human Genome Project has made great progress toward the goals of identifying all the 100,000 genes in
human DNA, determining the sequences of the 3 billion chemical base pairs that make up human DNA, storing
this information in databases, and developing tools for data analysis.
\begin{itemize}
\item
Each of these steps requires
sophisticated algorithms. Although the solutions to the various problems involved are beyond the scope of
this book, many methods to solve these biological problems use ideas from several of the chapters in this
book, thereby enabling scientists to accomplish tasks while using resources efficiently.

\item The savings are
in time, both human and machine, and in money, as more information can be extracted from laboratory
techniques.
\end{itemize}
\item
The Internet enables people all around the world to quickly access and retrieve large amounts of
information. With the aid of clever algorithms, sites on the Internet are able to manage and manipulate
this large volume of data. Examples of problems that make essential use of algorithms include finding good
routes on which the data will travel.

\item
Electronic commerce enables goods and services to be negotiated and exchanged electronically, and it
depends on the privacy of personal information such as credit card numbers, passwords, and bank
statements. The core technologies used in electronic commerce include public-key \hbox{cryptography} and digital
signatures (covered in Chapter 31), which are based on numerical algorithms and number theory. The savings are
in time, both human and machine, and in money, as more information can be extracted.

\end{itemize}

They have practical applications. Of the problems in the above list, finding the shortest path provides
the easiest examples. A transportation firm, such as a trucking or railroad company, has a financial
interest in finding shortest paths through a road or rail network because taking shorter paths results in
lower labor and fuel costs.

\subsection{Descriptive List}

Our first algorithm, insertion sort, solves the \textit{sorting problem} introduced in Chapter 1:
\begin{descriptive}
\item \head{Input} A sequence of $n$ numbers $\{a_{1},\ a_{2},\ \ldots,\ a_{n})$.

\item \head{Output} A permutation (reordering) $\{a_{1}',\ a_{2}',\ \ldots,\ a_{n}')$ of the input sequence such that
$a_{1}'\leq a_{2}'\geq\cdots\leq a_{n}'$.

\item \head{Divide}
 Divide the n-element sequence to be sorted into two subsequences of $n/2$ elements each.
\end{descriptive}

The numbers that we wish to sort are also known as the \textit{keys}. Although conceptually we are sorting
a sequence, the input comes to us in the form of an array with $n$ elements.

\subsection{List With Heading}

Consider the problem of adding two n-bit binary integers, stored in two n-element arrays $A$ and $B$. The
sum of the two integers should be stored in binary form in an $(n+1)$-element array $C$. State the problem
formally and write pseudocode for adding the two integers.

\listhead{Analyzing Algorithms}

\begin{enumerate}[2.]

\item
Otherwise, $u=2^{2^{k}}$ for some integer $k\geq 1$, so that $u\geq 4$. In addition to the universe size
$u$, the data structure \textit{pro to-vEB} $(u)$ contains the following attributes, illustrated.

\item An array \textit{cluster} $[0.\ .\ \sqrt{u}-1]$ of $\sqrt{u}$ pointers, each to a \textit{pro to-vEB}
$(\sqrt{u})$ structure.

\end{enumerate}

\subsection{Notation}

\begin{notation}[MDA$(H)$]

\item[MDA$(H)$] Maximum domain of attraction of the value distribution $H$

\item[N$(\mu,\sigma^2)$] Gaussian (normal) distribution with mean $\mu$, variance $\sigma^2$

\item[N$(\mu, \Sigma)$] Multivariate Gaussian (normal) distribution with mean vector $\mu$ and covariance
matrix $\Sigma$
\end{notation}

\begin{figure}[b!]
\includegraphics[scale=.85]{fig1}
\caption{Sorting a Hand of Cards Using Insertion Sort.\label{fig1}}%
\end{figure}%

\begin{sidewaysfigure}
 \includegraphics[width=0.9\textwidth]{image-1}\vspace*{-3pt}
\caption{Textual Entailment Framework, as Described. Amount of Time Worked Annually in 7 OECD Countries
Over the Period 1970--2011 (Total Number of Hours Worked During the Year Divided by the Average Number of
Persons of Working Age).} \label{fig:discourse-commits}
\end{sidewaysfigure}

%%%%
\subsection{The Evolution of Participation Rates}\label{ch01:sec1.3}

For each function $f(n)$ and time $t$ in the following table, determine the largest size $n$ of a problem
that can be solved in time $t$, assuming that the algorithm to solve the problem takes $f(n)$
microseconds.

When we evaluated this summation, we attained a bound of $\Theta(n^{2})$ on the worst-case running time of
the algorithm. This example illustrates why you should know how to manipulate and bound summations. In
this equation, the $\Theta$-notation on the left-hand side applies to the variable $k$, but on the
right-hand side, it applies to $n$. We can also apply such manipulations to infinite convergent series.

\section{Arithmetic Series}

\subsection{Extracts}

\begin{extract}
The Human Genome Project has made great progress toward the goals of identifying all the 100,000 genes in
human DNA, determining the sequences of the 3 billion chemical base pairs that make up human DNA, storing
this information in databases, and developing tools for data analysis. Each of these steps requires
sophisticated algorithms. Although the solutions to the various problems involved are beyond the scope of
this book, many methods to solve these biological problems use ideas from several of the chapters in this
book, thereby enabling scientists to accomplish tasks while using resources efficiently. The savings are
in time, and in money, as more information can be extracted from laboratory techniques in
Figures~\ref{fig1} and \ref{fig:discourse-commits}.
\begin{extractlist}

\item They have many candidate solutions, the overwhelming majority of which do not solve the problem at hand.
Finding one that does, or one that is ``best,'' can present quite a challenge.

\item
They have practical applications. Of the problems in the above list, finding the shortest path provides
the easiest examples. A transportation firm, such as a trucking or railroad company, has a financial
interest in finding shortest paths through a road or rail network because taking shorter paths results in
lower labor and fuel costs. Or a routing node on the Internet may need to find the shortest path through
the network in order to route a message quickly. Or a person wishing to drive from New York to Boston may
want to find driving directions from an appropriate Web site, or she may use her GPS while driving.

\item No black-heights in the tree have changed.

\item No red nodes have been made adjacent. Because $y$ takes $z$'s place in the tree, along with $z$'s color,
we cannot have two adjacent red nodes at $y$'s new position in the tree. In addition, if $y$ was not $z$'s
right child, then $y$'s original right child $x$ replaces $y$ in the tree. If $y$ is red, then $x$ must be
black, and so replacing $y$ by $x$ cannot cause two red nodes to become adjacent.

\end{extractlist}

Sorting is by no means the only computational problem for which algorithms have been developed. (You
probably suspected as much when you saw the size of this book.) Practical applications of algorithms are
ubiquitous and include the following examples:
\begin{unlist}
\item An array \textit{cluster} $[0\ldots \sqrt{u}-1]$ of $\sqrt{u}$ pointers, each to a \textit{pro to-vEB}
$(\sqrt{u})$ structure.

\item If $u=2$, then it is the base size, and it contains an array $A[0\ldots 1]$ of two bits.

\item Since $y$ could not have been the root if it was red, the root remains black.

\end{unlist}

With the aid of clever algorithms, sites on the Internet are able to manage and manipulate this large
volume of data. Examples of problems that make essential use of algorithms include finding good routes on
which the data will travel (techniques for solving such problems appear in Chapter 24), and using a search
engine to quickly find pages on which particular information resides (related techniques are in Chapters
11 and 32). This book also contains several data structures. For example, we might need to sort a sequence
of numbers into nondecreasing order. This problem arises frequently in practice and provides fertile
ground for introducing many standard design techniques and analysis tools. Here is how we formally define
the \textit{sorting problem}:
\begin{itemize}
\item
The Internet enables people all around the world to quickly access and retrieve large amounts of
information. With the aid of clever algorithms, sites on the Internet are able to manage and manipulate
this large volume of data. Examples of problems that make essential use of algorithms include finding good
routes on which the data will travel (techniques for solving such problems appear in Chapter 24), and
using a search engine to quickly find pages on which particular information resides (related techniques
are in Chapters 11 and 32). This book also contains several data structures. A {\it data structure} is a
way to store and organize data in order to facilitate access and modifications. No single data structure
works well for all purposes.

\item
Electronic commerce enables goods and services to be negotiated and exchanged electronically, and it
depends on the privacy of personal information such as credit card numbers, passwords, and bank
statements. The core technologies used in electronic commerce include public-key cryptography and digital
signatures (covered in Chapter 31), which are based on numerical algorithms and number theory.
\end{itemize}

The running time of the algorithm is the sum of running times for each statement executed; a statement
that takes $c_{i}$ steps to execute and executes $n$ times will contribute $c_{i}n$ to the total running
time.${}^{\text{6}}$ To compute $T(n)$, the running time of INSERTION-SORT on an input of $n$ values, we
sum the products of the \textit{cost} and \textit{times} columns, obtaining
\[
T(n) = c_{1}n+c_{2}(n-1)+c_{4}(n-1)+c_{5} \sum_{j=2}^{n}t_{j}+c_{6}\sum_{j=2}^{n}(t_{j}-1)
\]

The key operation of the merge sort algorithm is the merging of two sorted sequences in the ``combine''
step. We merge by calling an auxiliary procedure MERGE. $\mathrm{A};p, q, r$), where $A$ is an array and
$p,\ q$, and $r$ are indices into the array such that $p\geq q<r$. The procedure assumes that the
subarrays $A[p.\ .q]$ and $A[q+1.\ .r]$ are in sorted order. It {\it merges} them to form a single sorted
subarray that replaces the current subarray $A[p.\ .r]$.

Even for inputs of a given size, an algorithm`s running time may depend on \textit{which} input of that
size is given. For example, in INSERTION-SORT, the best case occurs if the array is already sorted. For
each $j=2,3, \ldots,\ n$, we then find that $A[i]\leq{}$\textit{key} in line 5 when $i$ has its initial
value of $j-1$. Thus $t_{j}=1$ for $j=2,3, \ldots,\ n$, and the best-case running time is
\begin{numsentence}
\item It is true prior to the first iteration of the loop.

\item If it is true before an iteration of the loop, it remains true before the
next iteration.

\item Apply these principles using data and programs that allowing us to
replicate the main results of the paper of the estimation of labor supply elasticities.

\item
Come up with a real-world problem in which only the best solution will do. Then come up with one in which
a solution that is ``approximately'' the best is good enough.
\end{numsentence}

When the loop is a for loop, the moment at which we check the loop invariant just prior to the first
iteration is immediately after the initial assignment to the loop-counter variable and just before the
first test in the loop header. In the case of INSERTION-SORT, this time is after assigning 2 to the
variable $j$ but before the first test of whether $j\leq A$. \textit{length}. The recursion ``bottoms
out'' when the sequence to be sorted has length 1, in which case there is no work to be done, since every
sequence of length 1 is already in sorted order.

\begin{unnumsentence}
\item No black-heights in the tree have changed  in Table~\ref{time-use}.

\item No red nodes have been made adjacent. Because $y$ takes $z$'s place in the
tree, along with $z$'s color, we cannot have two adjacent red nodes at $y$'s new position in the tree. In
addition, if $y$ was not $z$'s right child, then $y$'s original right child $x$ replaces $y$ in the tree.
If $y$ is red, then $x$ must be black, and so replacing $y$ by $x$ cannot cause two red nodes to become
adjacent.

\end{unnumsentence}

The element $x$, where $0\leq x<u$, is recursively stored in the cluster numbered high $(x)$ as element
low $(x)$ within that cluster.

\end{extract}

\begin{table}[!t]
\processtable{Average Minutes Per Day by Activity and  Employment Status  in the United States in
2003--2006.\label{time-use}}{
\begin{tabular*}{26pc}{@{\extracolsep{\fill}}lcc@{}}
\Toprule & \multicolumn{1}{c}{Employed} & {Unemployed} \\
\Midrule
{Sleep} &  496 &  555 \\
%\Midrule
{Personal care and eating} & 110 & 197 \\
%\Midrule
{Home production, shopping, care of others} & 158 & 254 \\
%\Midrule
{Leisure, travel, sports, and socializing} &  320 & 442\\
%\Midrule
{Work} &  325 & 210 \\
%\Midrule
{Job search} & 321 & 323 \\
\Botrule
\end{tabular*}}{Note: This is where authors provide additional information about
      the data, including whatever notes are needed.
}\vspace*{-3pt}
\end{table}%

Even for inputs of a given size, an algorithm`s running time may depend on \textit{which} input of that
size is given. For example, in INSERTION-SORT, the best case occurs if the array is already sorted. For
each $j=2,3,\ \ldots,\ n$, we then find that $A[i]\leq$ \textit{key} in line 5 when $i$ has its initial
value of $j-1$. Thus $t_{j}=1$ for $j=2,3,\ \ldots,\ n$, and the best-case running time is
\begin{align*}
T(n) &= c_{1}n+c_{2}(n-1)+c_{4}(n-1)+c_{5}(n-1)+c_{8}(n-1)\\
&= (c_{1}+c_{2}+c_{4}+c_{5}+c_{8})n-(c_{2}+c_{4}+c_{5}+c_{8}).
\end{align*}
We can express this running time as $an+b$ for \textit{constants} $a$ and $b$ that depend on the statement
costs $c_{i}$; it is thus a \textit{linear function} of $n$.

If the array is in reverse sorted order--that is, in decreasing order--the worst case results. We must
compare each element $A[j]$ with each element in the entire sorted subarray $A[1\ldots j-1]$, and so
$t_{j}=j$ for $j=2,3, \ldots,\ n$. Noting that
\[
\sum_{j=2}^{n}j=\frac{n(n+1)}{2}-1
\]
and
\[
\sum_{j=2}^{n}(j-1)=\frac{n(n-1)}{2}
\]
(see Appendix A for a review of how to solve these summations), we find that in the worst case, the
running time of INSERTIONSORT is
\begin{align*}
T(n) &= c_{1}n+c_{2}(n-1)+c_{4}(n-1)+c_{5}\left(\frac{n(n+1)}{2}-1\right)\\
&\quad+c_{6}\left(\frac{n(n-1)}{2}\right) +C_{7}\left(\frac{n(n-1)}{2}\right) +c_{8}(n-1)\\
&= \left(\frac{c_{5}}{2}+\frac{c_{6}}{2}+\frac{c_{7}}{2}\right)
n^{2}+\left(c_{1}+c_{2}+c_{4}+\frac{c_{5}}{2}-\frac{c_{6}}{2}-\frac{c_{7}}{2}+c_{8}\right)n\\
&\quad-(c_{2}+c_{4}+c_{5}+c_{8}).
\end{align*}
We can express this worst-case running time as $an^{2}+bn+c$ for constants $a,\ b$, and $c$ that again
depend on the statement costs $c_{i}$; it is thus a \textit{quadratic function} of $n$.

\begin{lstlisting}[language=C]
void initializeMyStruct( myStruct* ms ) {
   static int modulo3 = 0;
   if (modulo3 == 2) {
       modulo3 = 0;
       if (checkDay('tue')) {
           ms->a = 1;
       }
   } else {
       modulo3++;
       ms->a = 0;
   }
}
\end{lstlisting}

An important notational convention we shall adopt is that \textit{logarithm functions will apply only to
the next term in the formula}, so that $\mathrm{lg} n+k$ will mean $\mathrm{lg} n$) $+k$ and not
$\mathrm{lg}(n+k)$. If we hold $b>1$ constant, then for $n>0$, the function $\log_{b}n$ is strictly
increasing.

\subsection{Dialogue}

Given a well-defined notion of the remainder of one integer when divided by another, it is convenient to
provide special notation to indicate equality of remainders. If $(a\ \mathrm{mod}\ n)=(b\mathrm{mod}\ n)$,
we write $a\equiv b(\mathrm{mod}\ n)$ and say that $a$ is \textit{equivalent} to $b$, modulo $n$. In other
words, $a\equiv b(\mathrm{mod}\ n)$ if $a$ and $b$ have the same remainder when divided by $n$.
Equivalently, $a\equiv b(\mathrm{mod}\ n)$ if and only if $n$ is a divisor of $b-a$. We write $a\not\equiv
b(\mathrm{mod}\ n)$ if $a$ is not equivalent to $b$, modulo $n$.
   \begin{dialogue}
     \speak{Vladimir} Whare are all these corpses from?
     \speak{Estragon} These skeletons.
     \speak{Vladimir} A charnel-house!  A charnel-house!
     \speak{Estragon} I'm in hell!
     \speak{Vladimir} Where were you?
     \speak{Estragon} They're coming there too!
     \speak{Vladimir}  We're surrounded! makes a rush  towards back.  Imbecile!  There's no way out there.
                      There!  Not a soul in sight!  Off you go!  Quick!
   \end{dialogue}

For any integer $a$ and any positive integer $n$, the value $a \mathrm{mod}\ n$ is the \textit{remainder}
(or \textit{residue}) of the quotient $a/n$. We can extend our notation to the case of two parameters $n$
and $m$ that can go to infinity independently at different in rates.

\subsection{In Text Table}

For any real number $x$, we denote the greatest integer less than or equal to $x$ by $\lfloor x\rfloor$
(read the floor of $x$) and the least integer greater than or equal to $x$ by $\lceil x\rceil$ (read ``the
ceiling of $x$). For all real $x$,
\begin{table}[!h]
\unprocesstable{}{
\begin{tabular*}{26pc}{@{\extracolsep{\fill}}llccc@{}}
\Toprule
        Stubhead & \( df \) & \( f \) & \( \eta \) & \( p \) \\
\Midrule
                 &     \multicolumn{4}{c}{Spanning text}     \\%[-3pt]
%&\multicolumn{4}{c@{}}{\hrulefill}\\
        Row 1    & 1        & 0.67    & 0.55       & 0.41    \\
        Row 2    & 2        & 0.02    & 0.01       & 0.39    \\
        Row 3    & 3        & 0.15    & 0.33       & 0.34    \\
        Row 4    & 4        & 1.00    & 0.76       & 0.54    \\
\Botrule
     \end{tabular*}}{}
\end{table}

We can extend our notation to the case of two parameters $n$ and $m$ that can go to infinity independently
at different rates. For a given function $g(n,\ m)$, we denote by $O(g(n,\ m))$ the set of functions. For
any integer $a$ and any positive integer $n$, the value $a \mathrm{mod}\ n$ is the \textit{remainder} (or
\textit{residue}) of the quotient $a/n$. We can extend our notation to the case of two parameters $n$ and
$m$ that can go to infinity independently at different in rates.

\boxhead{Common Functions}

\begin{mdframed}
This section reviews some standard mathematical functions and notations and explores the relationships
among them. It also illustrates the use of the asymptotic notations.

As an example of the application of this theorem, our proof $an^{2}+bn+c= \Theta(n^{2})$ for any constants
$a,\ b$, and $c$, where $a>0$, immediately implies that $an^{2}+bn+c=\Omega(n^{2})$ and
$an^{2}+bn+c=O(n^{2})$. In practice, rather than using Theorem 3.1 to obtain asymptotic upper and lower
bounds from asymptotically tight bounds, as we did for this example, we usually use it to prove
asymptotically tight bounds from asymptotic upper and lower bounds. We have already seen how asymptotic notation can be used within mathematical formulas. For example, in
introducing O-notation.

\begin{extract}
Although the solutions to the various problems involved are beyond the scope of this book, many methods to
solve these biological problems use ideas from several of the chapters in this book, thereby enabling
scientists to accomplish tasks while using resources efficiently. The savings are in time, both human and
machine, and in money, as more information can be extracted from laboratory techniques.
\end{extract}

The running time of insertion sort therefore belongs to both $\Omega(n)$ and $O(n^{2})$, since it falls
anywhere between a linear function of $n$ and a quadratic function of $n$. Moreover, these bounds are
asymptotically as tight as possible: for instance, the running time of insertion sort is not
$\Omega(n^{2})$, since there exists an input for which insertion sort runs in $\Theta(n)$ time (e.g., when
the input is already sorted). It is not contradictory, however, to say that the \textit{worst-case}
running time of insertion sort is $\Omega(n^{2})$, since there exists an input that causes the algorithm
to take $\Omega(n^{2})$ time.

We have already seen how asymptotic notation can be used within mathematical formulas. For example, in
introducing O-notation, we wrote $n=O(n^{2})$. We might also write $2n^{2}+3n+1=2n^{2}+\Theta(n)$. How do
we interpret such formulas?
\[
T(n) = c_{1}n+c_{2}(n-1)+c_{4}(n-1)+c_{5} \sum_{j=2}^{n}t_{j}+c_{6}\sum_{j=2}^{n}(t_{j}-1)
\]

When the asymptotic notation stands alone (that is, not within a larger formula) on the right-hand side of
an equation (or inequality), as in $n=O(n^{2})$, we have already defined the equal sign to mean set
membership: $n\in O(n^{2})$. In general, however, when asymptotic notation appears in a formula, we
interpret it as standing for some anonymous function that we do not care to name. For example, the formula
$2n^{2}+3n+1=2n^{2}+\Theta(n)$ means that $2n^{2}+3n+1=2n^{2}+f(n)$, where $f(n)$ is some function in the
set $\Theta(n)$.

Using asymptotic notation in this manner can help eliminate inessential detail and clutter in an equation.
For example, in Chapter 2 we expressed the worst-case running time of merge sort as the recurrence
$T(n)=2T(n/2)+\Theta(n)$.

\section{Insertion Sort}

The numbers that we wish to sort are also known as the \textit{keys}. Although conceptually we are sorting
a sequence, the input comes to us in the form of an array with $n$ elements.

In this book, we shall typically describe algorithms as programs written in a \textit{pseudocode} that is
similar in many respects to $\mathrm{C},\ \mathrm{C}++$, Java, Python, or Pascal. If you have been
introduced to any of these languages, you should have little trouble reading our algorithms. What
separates pseudocode from ``real'' code is that in pseudocode, we employ whatever expressive method is
most clear and concise to specify a given algorithm.

\subsection{The Divide-and-Conquer Approach}

Many useful algorithms are \textit{recursive} in structure: to solve a given problem, they call themselves
recursively one or more times to deal with closely related subproblems. These algorithms typically follow
a \textit{divide-and-conquer} approach: they break the problem into several subproblems that are similar
to the original problem but smaller in size, solve the subproblems recursively, and then combine these
solutions to create a solution to the original problem.

\subsubsection{Order of Growth}

We used some simplifying abstractions to ease our analysis of the INSERTIONSORT procedure. First, we
ignored the actual cost of each statement, using the constants $c_{i}$ to represent these costs. Then, we
observed that even these constants give us more detail than we really need: we expressed the worst-case
running time as $an^{2}+bn+c$ for some constants $a,\ b$, and $c$ that depend on the statement costs
$c_{i}$. We thus ignored not only the actual statement costs, but also the abstract costs $c_{i}$.

We shall now make one more simplifying abstraction: it is the \textit{rate of growth}, or \textit{order of
growth}, of the running time that really interests us. We therefore consider only the leading term of a
formula $(\mathrm{e}.\mathrm{g}.,\ an^{2})$, since the lower-order terms are relatively insignificant for
large values of $n$. We also ignore the leading term's constant coefficient, since constant factors are
less significant than the rate of growth in determining computational efficiency for large inputs. %For
%insertion sort, when we ignore the lower-order terms and the leading term's constant coefficient, we are
%left with the factor of $n^{2}$ from the leading term. We write that insertion sort has a worst-case
%running time of $\Theta(n^{2})$ (pronounced ``theta of n-squared``). We shall use $\Theta$-notation
%informally in this chapter, and we will define it precisely in Chapter 3.

\paragraph{Analysis of Insertion Sort}

Since the number atoms in the observable universe is estimated to be about $10^{80}$, which is much less
than $2^{65536}$, we rarely encounter an input size $n$ such that $\mathrm{lg}^{*}n>5$. Fibonacci numbers
are related to the \textit{golden ratio} $\phi$ and to its conjugate $\hat{\phi}$, which are the two roots
of the equation.

Although the pseudocode for MERGE-SORT works correctly when the number of elements is not even, our
recurrence-based analysis is simplified if we assume that the original problem size is a power of 2. Each
divide step then yields two subsequences of size exactly $n/2$.

\begin{theorem}\label{thm2}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{theorem}

In Chapter 4, we shall see that this assumption does not affect the order of growth of the solution to the
recurrence.

\begin{remark}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{remark}

Since the number atoms in the observable universe is estimated to be about $10^{80}$, which is much less
than $2^{65536}$, we rarely encounter an input size $n$ such that $\mathrm{lg}^{*}n>5$. Fibonacci numbers
are related to the \textit{golden ratio} $\phi$ and to its conjugate $\hat{\phi}$, which are the two roots
of the equation.

\end{mdframed}

The Human Genome Project has made great progress toward the goals of identifying all the 100,000 genes in
human DNA, determining the sequences of the 3 billion chemical base pairs that make up human DNA, storing
this information in databases, and developing tools for data analysis. Each of these steps requires
sophisticated algorithms. Although the solutions to the various problems involved are beyond the scope of
this book, many methods to solve these biological problems use ideas from several of the chapters in this
book, thereby enabling scientists to accomplish tasks while using resources efficiently. The savings are
in time, both human and machine, and in money, as more information can be extracted from laboratory
techniques.

\begin{table}[b!]\vspace*{-6pt}
\processtable{Racial unemployment differences among men aged 16 and older in the United States in
2012.\label{usgapl}} {\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc@{}}
\Toprule
& \multicolumn{2}{c}{Unemployment Rate} \\
\cline{2-3}
& Annual Average & Ratio with Whites\\
\Midrule
Whites & 7.4 & 1.0 \\
Black or African American & 15\phantom{00} & 2.0 \\
Asian & 5.8 & 0.8 \\
Hispanic or Latino ethnicity & 9.9 & 1.4 \\
\Botrule
\end{tabular*}}{Source: Bureau of Labor Statistics, Labor force statistics from the current population survey, annual averages,
table 24.}
\end{table}%

This section reviews some standard mathematical functions and notations and explores the relationships
among them. It also illustrates the use of the asymptotic notations.

As an example of the application of this theorem, our proof $an^{2}+bn+c= \Theta(n^{2})$ for any constants
$a,\ b$, and $c$, where $a>0$, immediately implies that $an^{2}+bn+c=\Omega(n^{2})$ and
$an^{2}+bn+c=O(n^{2})$. In practice, rather than using Theorem 3.1 to obtain \hbox{asymptotic} upper and
lower bounds from \hbox{asymptotically} tight bounds, as we did for this example, we usually use it to
prove asymptotically tight bounds from asymptotic upper and lower bounds.
When we say that the \textit{running time} (no modifier) of an algorithm is $\Omega(g(n))$, we mean that
\textit{no matter what particular input of size} $n$ \textit{is chosen for each value} $ofn$, the running
time on that input is at least a constant times $g(n)$, for sufficiently large $n$. Equivalently, we are
giving a lower bound on the best-case running time of an algorithm. For example, the best-case running
time of insertion sort is $\Omega(n)$, which implies that the running time of insertion sort is
$\Omega(n)$ in Figure~\ref{usgapl}.

The running time of insertion sort therefore belongs to both $\Omega(n)$ and $O(n^{2})$, since it falls
anywhere between a linear function of $n$ and a quadratic function of $n$. Moreover, these bounds are
asymptotically as tight as possible: for instance, the running time of insertion sort is not
$\Omega(n^{2})$, since there exists an input for which insertion sort runs in $\Theta(n)$ time (e.g., when
the input is already sorted). It is not contradictory, however, to say that the \textit{worst-case}
running time of insertion sort is $\Omega(n^{2})$, since there exists an input that causes the algorithm
to take $\Omega(n^{2})$ time.

We have already seen how asymptotic notation can be used within mathematical formulas. For example, in
introducing O-notation, we wrote $n=O(n^{2})$. We might also write $2n^{2}+3n+1=2n^{2}+\Theta(n)$. How do
we interpret such formulas?

When the asymptotic notation stands alone (that is, not within a larger formula) on the right-hand side of
an equation (or inequality), as in $n=O(n^{2})$, we have already defined the equal sign to mean set
membership: $n\in O(n^{2})$. In general, however, when asymptotic notation appears in a formula, we
interpret it as standing for some anonymous function that we do not care to name. For example, the formula
$2n^{2}+3n+1=2n^{2}+\Theta(n)$ means that $2n^{2}+3n+1=2n^{2}+f(n)$, where $f(n)$ is some function in the
set $\Theta(n)$. In this case, we let $f(n)=3n+1$, which indeed is in $\Theta(n)$.

\backmatter

\begin{exercises}

\begin{enumerate}

\item Express the function $n^{3}/1000-100n^{2}-100n+3$ in terms of O-notation.

\item  Consider sorting $n$ numbers stored in array $A$ by first finding the smallest element of $A$ and
exchanging it with the element in $A[1]$. Then find the second smallest element of $A$, and exchange it
with $A[2]$. Continue in this manner for the first $n-1$ elements of $A$. Write pseudocode for this
algorithm, which is known as \textit{selection sort}. What loop invariant does this algorithm maintain?
Why does it need to run for only the first $n-1$ elements, rather than for all $n$ elements? Give the
best-case and worst-case running times of selection sort in O-notation.
\begin{enumerate}

\item Consider linear search again (see exercise). How many elements of the input sequence need to be
checked on the average, assuming that the element being searched for is equally likely to be any element
in the array? How about in the worst case? What are the average-case and worst-case running times of
linear search in $\Theta$-notation? Justify your answers.

\begin{enumerate}

\item How can we modify almost any algorithm to have a good best-case running time?

\item
Rewrite the MERGE procedure so that it does not use sentinels, instead stopping once either array $L$ or
$R$ has had all its elements copied back to $A$ and then copying the remainder of the other array back
into $A$.

\item
Let $f(n)$ and $g(n)$ be asymptotically nonnegative functions. Using the basic definition of
$\Theta$-notation, prove that $\displaystyle \max(f(n),\ g(n))=\Theta(f(n)+g(n))$.

\end{enumerate}

\item
Show that for any real constants $a$ and $b$, where $b>0$, $(n+a)^{b}=\Theta(n^{b})$.

\item
Show that if $f(n)$ and $g(n)$ are monotonically increasing functions, then so are the functions
$f(n)+g(n)$ and $f(g(n))$, and if $f(n)$ and $g(n)$ are in addition nonnegative, then $f(n)\cdot g(n)$ is
monotonically increasing.

\end{enumerate}
\item
Is the function $\lceil \mathrm{lg}n\rceil$! polynomially bounded? Is the function $\lceil
\mathrm{lglg}n\rceil$ ! polynomially bounded?
\end{enumerate}

Prove by induction that the $i$ th Fibonacci number satisfies the equality $F_{i}=\displaystyle
\frac{\phi^{i}-\hat{\phi}^{i}}{\sqrt{5}}$ where $\phi$ is the golden ratio and $\hat{\phi}$ is its
conjugate.

\section{Insertion Sort}

The numbers that we wish to sort are also known as the \textit{keys}. Although conceptually we are sorting
a sequence, the input comes to us in the form of an array with $n$ elements.

In this book, we shall typically describe algorithms as programs written in a \textit{pseudocode} that is
similar in many respects to $\mathrm{C},\ \mathrm{C}++$, Java, Python, or Pascal. If you have been
introduced to any of these languages, you should have little trouble reading our algorithms. What
separates pseudocode from ``real'' code is that in pseudocode, we employ whatever expressive method is
most clear and concise to specify a given algorithm. Sometimes, the clearest method is English, so do not
be surprised if you come across an English phrase or sentence embedded within a section of ``real'' code.
Another difference between pseudocode and real code is that pseudocode is not typically concerned with
issues of software engineering. Issues data abstraction, modularity, and error handling are often ignored
in order to convey the essence of the algorithm more concisely.

\subsection{The Divide-and-Conquer Approach}

Many useful algorithms are \textit{recursive} in structure: to solve a given problem, they call themselves
recursively one or more times to deal with closely related subproblems. These algorithms typically follow
a \textit{divide-and-conquer} approach: they break the problem into several subproblems that are similar
to the original problem but smaller in size, solve the subproblems recursively, and then combine these
solutions to create a solution to the original problem.

\subsubsection{Order of Growth}

We used some simplifying abstractions to ease our analysis of the INSERTIONSORT procedure. First, we
ignored the actual cost of each statement, using the constants $c_{i}$ to represent these costs. Then, we
observed that even these constants give us more detail than we really need: we expressed the worst-case
running time as $an^{2}+bn+c$ for some constants $a,\ b$, and $c$ that depend on the statement costs
$c_{i}$. We thus ignored not only the actual statement costs, but also the abstract costs $c_{i}$.

We shall now make one more simplifying abstraction: it is the \textit{rate of growth}, or \textit{order of
growth}, of the running time that really interests us. We therefore consider only the leading term of a
formula $(\mathrm{e}.\mathrm{g}.,\ an^{2})$, since the lower-order terms are relatively insignificant for
large values of $n$. We also ignore the leading term's constant coefficient, since constant factors are
less significant than the rate of growth in determining computational efficiency for large inputs. For
insertion sort, when we ignore the lower-order terms and the leading term's constant coefficient, we are
left with the factor of $n^{2}$ from the leading term. We write that insertion sort has a worst-case
running time of $\Theta(n^{2})$ (pronounced ``theta of n-squared``). We shall use $\Theta$-notation
informally in this chapter, and we will define it precisely in Chapter 3.

\paragraph{Analysis of insertion sort}

A better understanding of labor supply behaviors is also key for the design of tax systems in general. The
more sensitive the labor supply to the net wage, the lower the optimal tax rate, because high tax rates
will then tend to shrink the source of taxable income. This sensitivity, also called elasticity, might
vary substantially across gender, age and income groups. Another motivation is the need to analyse
fluctuations of employment over the economic cycle, which depend on how labor supply reacts to changes in
wages. It is assumed that the individual is ready to sacrifice less and less consumption for an extra hour
of leisure when the amount of time dedicated to leisure rises. This property signifies that the marginal
rate of substitution between consumption and leisure diminishes with leisure time, or again that the
indifference curves are convex, which is equivalent to the hypothesis of the quasi-concavity of the
utility function (the relation between the shape of the indifference curves and the utility function is
studied in appendix 1 at the end of this chapter). Note that for a married or cohabiting person, a part of
the income of her partner is capable of being integrated into this set. Thus the budget constraint of the
agent takes the form.

\subparagraph{Analysis of merge sort}

Although the pseudocode for MERGE-SORT works correctly when the number of elements is not even, our
recurrence-based analysis is simplified if we assume that the original problem size is a power of 2. Each
divide step then yields two subsequences of size exactly $n/2$. In Chapter 4, we shall see that this
assumption does not affect the order of growth of the solution to the recurrence.

The notations we use to describe the asymptotic running time of an algorithm are defined in terms of
functions whose domains are the set of natural numbers $\mathbb{N}=\{0,1,2,\ \ldots\}$. Such notations are
convenient for describing the worst-case running-time function $T(n)$, which usually is defined only on
integer input sizes. We sometimes find it convenient, however, to \textit{abuse} asymptotic notation in a
variety of ways. For example, we might extend the notation to the domain of real numbers or,
alternatively, restrict it to a subset of the natural numbers. We should make sure, however, to understand
the precise meaning of the notation so that when we abuse, we do not \textit{misuse} it. This section
defines the basic asymptotic notations and also introduces some common abuses.

\begin{theorem}\label{thm1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{theorem}

Indentation indicates block structure. For example, the body of the for loop that begins on line 1
consists of lines 2-8, and the body of the while loop that begins on line 5 contains lines 6-7 but not
line 8. Our indentation style applies to if-else statements as well. Using indentation instead of
conventional indicators of block structure, such as begin and end statements, greatly reduces clutter
while preserving, or even enhancing, clarity.

\begin{remark}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{remark}

The looping constructs while, for, and repeat-until and the if-else conditional construct have
interpretations similar to those in $\mathrm{C},\ \mathrm{C}++$, Java, Python, and Pascal. In this book,
the loop counter retains its value after exiting the loop, unlike some situations that arise in
$\mathrm{C}++$, Java, and Pascal. Thus, immediately after a for loop, the loop counter`s value is the
value that first exceeded the for loop bound. We used this property in our correctness argument for
insertion sort. The for loop header in line 1 is for $j=2$ to \textit{A. length}, and so when this loop
terminates, $j=A. length+1$ (or, equivalently, $j=n+1$, since $n=A$. \textit{length}). We use the keyword
to when a for loop increments its loop counter in each iteration, and we use the keyword downto when a for
loop decrements its loop counter. When the loop counter changes by an amount greater than 1, the amount of
change follows the optional keyword by.

\end{exercises}

\setcounter{chapter}{0}
\begin{appendices}
\chapter{Summations}

When an algorithm contains an iterative control construct such as a while or for loop, we can express its
running time as the sum of the times spent on each execution of the body of the loop. For example, we
found in Section 2.2 that the jth iteration of insertion sort took time proportional to $j$ in the worst
case. By adding up the time spent on each iteration, we obtained the summation (or series)
\[
\sum_{j=2}^{n}j.
\]

When we evaluated this summation, we attained a bound of $\Theta(n^{2})$ on the worst-case running time of
the algorithm. This example illustrates why you should know how to manipulate and bound summations.

\section{Summation Formulas and Properties}

Given a sequence $a_{1}, a_{2}, \ldots$ , $a_{n}$ of numbers, where $n$ is a nonnegative integer, we can
write the finite sum $a_{1}+a_{2}+\cdots+a_{n}$ as
\[
\sum_{k=1}^{n}a_k.
\]
If $n=0$, the value of the summation is defined to be $0$. The value of a finite series is always well
defined, and we can add its terms in any order.

Given an infinite sequence $a_{1}, a_{2}, \ldots$ of numbers, we can write the infinite sum
$a_{1}+a_{2}+\cdots$ as
\[
\sum_{k=1}^{\infty}a_k,
\]
which we interpret to mean
\[
\lim_{n\to \infty}\sum_{k=1}^{\infty}a_k.
\]
If the limit does not exist, the series \textit{diverges}; otherwise, it \textit{converges}. The terms of
a convergent series cannot always be added in any order. We can, however, rearrange the terms of an
\textit{absolutely convergent series}, that is, a series $ \sum_{k=1}^{\infty}a_{k}$ for which the series
$\sum_{k=1}^{\infty}|a_{k}|$ also converges.

\subsection{Linearity}

For any real number $c$ and any finite sequences $a_{1}, a_{2}, \ldots$ , $a_{n}$ and $b_{1}, b_{2},
\ldots$ , $b_{n}$,
\[
\sum_{k=1}^{n} (ca_k + b_k) = c \sum_{k=1}^{n} a_k + \sum_{k=1}^{n} b_k.
\]
The linearity property also applies to infinite convergent series.

We can exploit the linearity property to manipulate summations incorporating asymptotic notation. For
example,
\[
\sum_{k=1}^{n} \Theta (f(k)) =\Theta \left( \sum_{k=1}^{n} f(k)\right).
\]
In this equation, the $\Theta$-notation on the left-hand side applies to the variable $k$, but on the
right-hand side, it applies to $n$. We can also apply such manipulations to infinite convergent series.

\subsubsection{Arithmetic Series}

The summation
\[
\sum_{k=1}^{n} k = 1+ 2+\cdots + n.
\]
is an \textit{arithmetic series} and has the value
\begin{align}
\sum_{k=1}^{n} k & = \frac{1}{2} (n+1)\\
=\Theta (n^2).
\end{align}

%%%%%

\paragraph{Sums of squares and cubes}

We have the following summations of squares and cubes:
\begin{align}
\sum_{k=0}^{n} k^2  &= \frac{n(n+1)(2n+1)}{6},\\
\sum_{k=0}^{n} k^3  &=\frac{n^2(n+1)^2}{4}.
\end{align}

\subparagraph*{Geometric series}

For real $x \neq 1$, the summation
\[
\sum_{k=0}^{n} x^k= 1 + x + x^2 +\cdots+ x^n
\]
Because we assume that $0^0 = 1$, these formulas apply even when $x = 0$.

%%%%

Indentation indicates block structure. For example, the body of the for loop that begins on line 1
consists of lines 2-8, and the body of the while loop that begins on line 5 contains lines 6-7 but not
line 8. Our indentation style applies to if-else statements as well. Using indentation instead of
conventional indicators of block structure, such as begin and end statements, greatly reduces clutter
while preserving, or even enhancing, clarity.

\begin{remark}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{remark}

The looping constructs while, for, and repeat-until and the if-else conditional construct have
interpretations similar to those in $\mathrm{C},\ \mathrm{C}++$, Java, Python, and Pascal.${}^{\text{4}}$
In this book, the loop counter retains its value after exiting the loop, unlike some situations that arise
in $\mathrm{C}++$, Java, and Pascal. Thus, immediately after a for loop, the loop counter`s value is the
value that first exceeded the for loop bound.

\begin{theorem}\label{thm1}
For any two functions $f(n)$ and $g(n)$, we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and
$f(n)=\Omega(g(n))$. As an example of the application of this theorem, our proof that $an^{2}+bn+c=
\Theta(n^{2})$ for any constants $a, b$, and $c$, where $a>0$, immediately implies that
$an^{2}+bn+c=\Omega(n^{2})$ and $an^{2}+bn+c=O(n^{2})$.
\end{theorem}

In practice, rather than using Theorem~\ref{thm1} to obtain asymptotic upper and lower bounds from
asymptotically tight bounds, as we did for this example, we usually use it to prove asymptotically tight
bounds from asymptotic upper and lower bounds in \ref{cor1}.

\subsection{Extracts}

The Human Genome Project has made great progress toward the goals of identifying all the 100,000 genes in
human DNA, determining the sequences of the 3 billion chemical base pairs that make up human DNA, storing
this information in databases, and developing tools for data analysis. Each of these steps requires
sophisticated algorithms. The Human Genome Project has made great progress toward the goals of identifying all the 100,000 genes in
human DNA, determining the sequences of the 3 billion chemical base pairs that make up human DNA, storing
this information in databases, and developing tools for data analysis. Each of these steps requires
sophisticated algorithms.

\begin{extract}
Although the solutions to the various problems involved are beyond the scope of this book, many methods to
solve these biological problems use ideas from several of the chapters in this book, thereby enabling
scientists to accomplish tasks while using resources efficiently. The savings are in time, both human and
machine, and in money, as more information can be extracted from laboratory techniques.
\end{extract}
Indicate, for each pair of expressions $(A,\ B)$ in the table below, whether $A$ is $O$, $0$, $\Omega$,
$\omega$, or $\Theta$ of $B$. Assume that $k\leq 1$, $\epsilon>0$, and $c>1$ are constants. Your answer
should be in the form of the table with ``yes'' or ``no'' written in each box.

\subsection{The Divide-and-Conquer Approach}

Many useful algorithms are \textit{recursive} in structure: to solve a given problem, they call themselves
recursively one or more times to deal with closely related subproblems. These algorithms typically follow
a \textit{divide-and-conquer} approach: they break the problem into several subproblems that are similar
to the original problem but smaller in size, solve the subproblems recursively, and then combine these
solutions to create a solution to the original problem.

\subsection{Order of Growth}

We used some simplifying abstractions to ease our analysis of the INSERTIONSORT procedure. First, we
ignored the actual cost of each statement, using the constants $c_{i}$ to represent these costs. Then, we
observed that even these constants give us more detail than we really need: we expressed the worst-case
running time as $an^{2}+bn+c$ for some constants $a,\ b$, and $c$ that depend on the statement costs
$c_{i}$. We thus ignored not only the actual statement costs, but also the abstract costs $c_{i}$ to the
original problem.

\subsection{Approach}

Many useful algorithms are \textit{recursive} in structure: to solve a given problem, they call themselves
recursively one or more times to deal with closely related subproblems. These algorithms typically follow
a \textit{divide-and-conquer} approach: they break the problem into several subproblems that are similar
to the original problem but smaller in size, solve the subproblems recursively, and then combine these
solutions to create a solution to the original problem.

\subsection{Growth}

We used some simplifying abstractions to ease our analysis of the INSERTIONSORT procedure. First, we
ignored the actual cost of each statement, using the constants $c_{i}$ to represent these costs. Then, we
observed that even these constants give us more detail than we really need: we expressed the worst-case
running time as $an^{2}+bn+c$ for some constants $a,\ b$, and $c$ that depend on the statement costs
$c_{i}$. We thus ignored not only the actual statement costs, but also the abstract costs $c_{i}$ to the
original problem.

\subsection{Order of Growth}

We used some simplifying abstractions to ease our analysis of the INSERTIONSORT procedure. First, we
ignored the actual cost of each statement, using the constants $c_{i}$ to represent these costs. Then, we
observed that even these constants give us more detail than we really need: we expressed the worst-case
running time as $an^{2}+bn+c$ for some constants $a,\ b$, and $c$ that depend on the statement costs
$c_{i}$. We thus ignored not only the actual statement costs, but also the abstract costs $c_{i}$ to the
original problem.

\subsection{Approach}

Many useful algorithms are \textit{recursive} in structure: to solve a given problem, they call themselves
recursively one or more times to deal with closely related subproblems. These algorithms typically follow
a \textit{divide-and-conquer} approach: they break the problem into several subproblems that are similar
to the original problem but smaller in size, solve the subproblems recursively, and then combine these
solutions to create a solution to the original problem to the original problem.

\end{appendices}

%%
\chapter*{Notes}

\begin{notes}[10.]

\item
Knuth [209] traces the origin of the O-notation to a number-theory text by P. Bachmann in 1892. The
o-notation was invented by E. Landau in 1909 for his discussion of the distribution of prime numbers. The
$\Omega$ and $\Theta$ notations were advocated by Knuth [213] to correct the popular, but technically
sloppy, practice in the literature of using O-notation for both upper and lower bounds. Many people
continue to use the O-notation where the $\Theta$-notation is more technically precise. Further discussion
of the history and development of asymptotic notations appears in works by Knuth [209, 213] and Brassard
and Bratley [54].

\item
Not all authors define the asymptotic notations in the same way, although the various definitions agree in
most common situations. Some of the alternative definitions encompass functions that are not
asymptotically nonnegative, as long as their absolute values are appropriately bounded.

\item Knuth [211] provides an encyclopedic treatment of many sorting algorithms. His
comparison of sorting algorithms (page 381) includes exact step-counting analyses, like the one we
performed here for insertion sort. Knuthís discussion of insertion sort encompasses several variations of
the algorithm. The most important of these is Shellís sort, introduced by D. L. Shell, which uses
insertion sort on periodic subsequences of the input to produce a faster sorting algorithm in \cite{1}.

\item Not all authors define the asymptotic notations in the same way, although the
various definitions agree in most common situations. Some of the alternative definitions encompass
functions that are not asymptotically nonnegative \cite{2}, as long as their absolute values are
appropriately bounded.

\item Equation (3.20) is due to Robbins [297]. Other properties of elementary mathematical
functions can be found in any good mathematical reference, such as Abramowitz and Stegun [1] or Zwillinger
[362], or in a calculus book, such as \cite{4} or \cite{3}. Knuth [209] and Graham, Knuth, and Patashnik
[152] contain a wealth of material on discrete mathematics as used in computer science.

\item We use min-heaps to implement min-priority queues in Chapters 16, 23, and 24.
We also give an implementation with improved time bounds for certain operations in Chapter 19 and,
assuming that the keys are drawn from a bounded set of nonnegative integers, Chapter 20.

\item The quicksort procedure was invented by \cite{5}; Hoareís version appears in
Problem 7-1. The PARTITION procedure given in Section 7.1 is due to N. Lomuto. The analysis in Section 7.4
is due to \cite{10}. Sedgewick [305] and \cite{9} provide a good reference on the details of
implementation and how they matter.

\item The decision-tree model for studying comparison sorts was introduced by Ford
and Johnson [110]. Knuthís comprehensive treatise on sorting \cite{8} covers many variations on the
sorting problem, including the information-theoretic lower bound on the complexity of sorting given here.
Ben-Or [39] studied lower bounds for sorting using generalizations of the decision-tree model.

\item The origin of stacks and queues as data structures in computer science is unclear,
since corresponding notions already existed in mathematics and paper-based business practices before the
introduction of digital computers. Knuth [209] cites A. M. Turing for the development of stacks for
subroutine linkage in 1947.

\item Carter and Wegman introduced the notion of universal classes of hash functions
in 1979 [58].
\end{notes}

\chapter*{Glossary}

\begin{glossary}

\item[formula] A mathematical expression

\item[latex] Is a mark up language specially suited for scientific documents. Turing for the development of stacks for
subroutine linkage

\item[mathematics] Mathematics is what mathematicians do Is a mark up language specially suited for scientific documents. Turing for the development of stacks for
subroutine linkage

\end{glossary}

\begin{thebibliography}{11}

\bibitem{1} Milton Abramowitz and Irene A. Stegun, editors. Handbook of Mathematical Functions.
Dover, 1965.

\bibitem{2} G. M. AdelÅfson-VelÅfski.. and E. M. Landis. An algorithm for the organization of information.
Soviet Mathematics Doklady, 3(5):1259.1263, 1962.

\bibitem{3} Alok Aggarwal and Jeffrey Scott Vitter. The input/output complexity of sorting and related
problems. Communications of the ACM, 31(9):1116.1127, 1988.

\bibitem{4} Manindra Agrawal, Neeraj Kayal, and Nitin Saxena. PRIMES is in P. Annals of Mathematics,
160(2):781.793, 2004.

\bibitem{5} Alfred V. Aho, John E. Hopcroft, and Jeffrey D. Ullman. The Design and Analysis of
Computer Algorithms. Addison-Wesley, 1974.

\bibitem{6} Alfred V. Aho, John E. Hopcroft, and Jeffrey D. Ullman. Data Structures and Algorithms.
Addison-Wesley, 1983.

\bibitem{7} Ravindra K. Ahuja, Thomas L. Magnanti, and James B. Orlin. Network Flows: Theory,
Algorithms, and Applications. Prentice Hall, 1993.

\bibitem{8} Ravindra K. Ahuja, Kurt Mehlhorn, James B. Orlin, and Robert E. Tarjan. Faster algorithms
for the shortest path problem. Journal of the ACM, 37(2):213.223, 1990.

\bibitem{9} Ravindra K. Ahuja and James B. Orlin. A fast and simple algorithm for the maximum flow
problem. Operations Research, 37(5):748.759, 1989.

\bibitem{10} Ravindra K. Ahuja, James B. Orlin, and Robert E. Tarjan. Improved time bounds for the
maximum flow problem. SIAM Journal on Computing, 18(5):939.954, 1989.

\bibitem{11} MiklÅLos Ajtai, Nimrod Megiddo, and Orli Waarts. Improved algorithms and analysis for
secretary problems and generalizations. In Proceedings of the 36th Annual Symposium on
Foundations of Computer Science, pages 473.482, 1995.

\end{thebibliography}

\begin{theindex}

\item AA-tree, 338
\item abelian group, 940
\item ABOVE, 1024
\item absent child, 1178
\item absolutely convergent series, 1146
\item absorption laws for sets, 1160
\item abstract problem, 1054
\item acceptable pair of integers, 972
\item acceptance
\subitem by an algorithm, 1058
\subitem by a finite automaton, 996
\item accepting state, 995
\item accounting method, 456ñ459
\subitem for binary counters, 458
\subitem for dynamic tables, 465ñ466
\subitem for stack operations, 457ñ458, 458 ex.
\subitem minimum spanning, see minimum spanning tree
\subitem optimal binary search, 397ñ404, 413
\item Ackermannís function, 585
\item activity-selection problem, 415ñ422, 450

\indexspace

\item babyface, 602 ex.
\item back edge, 609, 613
\item back substitution, 817
\item BAD-SET-COVER-INSTANCE, 1122 ex.
\item BALANCE, 333 pr.
\item balanced search tree
\subitem AA-trees, 338
\subitem AVL trees, 333 pr., 337
\subitem B-trees, 484ñ504
\item base case, 65, 84
\item base, in DNA, 391
\item basic feasible solution, 866

\indexspace

\item cache hit, 449 pr.
\item cache miss, 449 pr.
\item cache obliviousness, 504
\item caching, off-line, 449 pr.
\item call
\subitem in a multithreaded computation, 776
\subitem of a subroutine, 23, 25 n.
\subitem by value, 21
\item call edge, 778
\item cancellation lemma, 907

\indexspace

\item unary, 1056
\item unbounded linear program, 851
\item unconditional branch instruction, 23
\item uncountable set, 1161
\item underdetermined system of linear equations, 814
\item underflow
\subitem of a queue, 234
\subitem of a stack, 233
\item undirected graph, 1168

\end{theindex}

\end{document}

### 2015-04-06 challenges.tex
missing

\section{Challenges and Limitations}

While noninterference is an elegant criterion that accounts simultaneously for both integrity and confidentiality threats, it suffers from several limitations and poses some nontrivial implementation challenges. We discuss these in turn.

\subsubsection{Strictness}

A straightforward fix for the defects exhibited by the code in Figure \ref{Fi:chIFS:example} is to downgrade the high value: Sanitize or validate it to prevent integrity attacks, or declassify it to prevent release of sensitive information. Notice, however, that the noninterference property --- in its current form --- is too strict to accomodate the notion of downgrading.

Indeed, in the practical setting of production sofrware, 
whose behavior is often guided by user input and whose output often reflects data originating from multiple users, it is extremely challenging to enforce full separation between high and low values. Doing so often comes at the cost of critical functionality.

As a natural example, an e-commerce site accepts login credentials from the user, which it utilizes to access the user's account, stored on a backend database, via security-sensitive calls that could potentiall lead to an SQL injection attack. For confidentiality, public data presented to the user, such as purchase recommendations, may be influenced by sensitive data such as the user's profile.

In response, a relaxed notion of noninterference has been proposed by XYZ, which augments the basic setting with the ability to transform high values into low values via downgrading. This enhancement permits the programmer to handle security-sensitive values as they propagate through the code. With this enhancement, we arrive from Table \ref{Ta:highlow} to Table \ref{Ta:highlowExtended}.

\begin{table}
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			& {\bf Integrity} & {\bf Confidentiality} \\ \hline
			{\bf High}	& Untrusted & Secret \\ \hline
			{\bf Low} & Trusted & Public \\ \hline		
			{\bf Downgrading} & Endorsement & Declassification \\
			\hline
		\end{tabular}
	\end{center}
	\caption{\label{Ta:highlowExtended}The meaning of high and low values with the addition of downgrading}
\end{table}

\subsubsection{Correctness of Downgrading}

The introduction of downgraders as a means to accommodate information flows that transition from high to low is necessary for practical adoption of the noninterference criterion. At the same time, however, it introduces a new challenge: How do we ascertain that the downgrading operation operates correctly?

If the downgrader is incorrect, or partial, then an attacker may still be able to exploit the system. We provide an example in Figure \ref{TateExample}. This code, taken from ...

\subsubsection{Covert Channels}

So far we have implicitly considered only direct flow of information. However, security threats occupy a broader scope. In particular, an attack could become possible via indirect mechanisms, such as exploiting time or heat measurements.

In general, we refer to mechanisms for information transfer as \emph{channels}. A \emph{channel} is a means, or mechanism, to signal information through a compute device. 
\emph{Covert channels}, or \emph{side channels}, are channels that exploit a mechanism whose primary purpose is not information transfer.

Covert channels divide into several categories, which we survey in turn.

\paragraph{Implicit flows.} The control-flow behavior of a program is a means to signal information. Branching decisions governed by high values may impact low variables, which creates a covert channel.

\begin{figure}
	\begin{lstlisting}
	t = h % 2;
	l = 0;
	if (t == 1) 
	l = 1;
	\end{lstlisting}
	\caption{\label{Fi:implicitFlow}Example of an implicit flow}
\end{figure}

To illustrate this, we refer to the code in Figure \ref{Fi:implicitFlow}. In this program, we assume that {\tt l} and {\tt h} are low and high variables, respectively. Though there is no explicit assignment of the high value pointed-to by {\tt h} to {\tt l}, this program is equivalent to the program
\begin{quote}
	{\tt l = h \% 2}
\end{quote} 
where there is direct flow from high to low.

\paragraph{Termination.} Another channel, similar to implicit flows, is termination. Whether or not a program terminates is a potential channel for release of sensitive information. *** Example ***

\paragraph{Timing.} Monitoring the execution time of a program discloses information about the values it operates on. In a seminal paper \cite{XXX},

\paragraph{Probabilistic.} Under the assumption that the attacker is able to execute the program multiple times, sensitive information is potentially leaked through stochastic program properties by changing the distribution of observable data.

\paragraph{Resource exhaustion.} Another channel is exhaustion of finite compute resources, such as memory or disk space.

\paragraph{Power.} Similarly to time measurement, the power consumed by the compute device throughout the program's execution is a potential hint as to which values are being processed.

\paragraph{Sensors.} Correlations between sensors can disclose to an attacker unintended information. A recent example is use of the gyroscope built into modern smartphones to recover the speaker's identity, and even parse speech.

### 2015-04-06 noninterference.tex
missing

\section{The Principle of Noninterference}

An elegant way of formalizing the requirement that the source data does not affect the sink's behavior is via the principle of noninterference. This principle --- first proposed by Goguen and Meseguer in 1982 \cite{GoguenMeseguer} --- formalizes the intuition that variation ``sensitive'' values should not cause variation of ``nonsensitive'' values.

Noninterference assumes an abstract model of the program as an input/output machine. Inputs and outputs are classified as either \emph{high} or \emph{low}.
%
In confidentiality, low is interpreted as low sensitivity. A low value is thus a value that is considered unclassified. Complementarily, a high value is considered classified (or sensitive). Integrity assigns the dual meaning to high and low. 
In integrity, low means trusted, whereas high means untrusted. Thanks to these dual interpretations, both integrity and confidentiality reduce to the requirement that there be no flow of high input values to the low part of the output.
%
These dual interpretations of high and low are summarized in Table \ref{Ta:highlow}.

\begin{table}
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			& {\bf Integrity} & {\bf Confidentiality} \\ \hline
			{\bf High}	& Untrusted & Secret \\ \hline
			{\bf Low} & Trusted & Public \\ 		
			\hline
		\end{tabular}
	\end{center}
	\caption{\label{Ta:highlow}The meaning of high and low values with the addition of downgrading}
\end{table}

With this background, we can give an informal statement of the property of noninterference. The property holds if and only if the restriction of the output to low values is the same for any pair of inputs that agree on the low values. That is, modifying the high values does not lead to any observable change in the system's output. In this sense, there is no interference between high and low values.

Formal statement of the noninterference principle requires some notation. Denote by $M$ a memory configuration, and let $M_L$ and $M_H$ denote the projection of $M$ to its low and high parts, respectively. We specialize the equality relation analogously: $M \equiv_{L} M'$ means equality over the low values between memory configurations $M$ and $M'$, and similarly, $M \equiv_{H} M'$ means equality over the high parts. Finally, let $P$ be the program, which we assume for simplicity to be deterministic, and $(P,M) \longrightarrow^{*} M'$ a complete execution of $P$ starting at memory configuration $M$ and terminating at configuration $M'$.

The definition of noninterference is then as follows:
$$
\begin{array}{rrrcl}
\forall M_1,M_2. &  				  & M_1 & \equiv_L & M_2 \\
&  \wedge & (P,M_1) & \longrightarrow^{*} & M_1' \\ 
&  \wedge & (P,M_2) & \longrightarrow^{*} & M_2' \\
& \Longrightarrow & M_1'    & \equiv_L & M_2'  \\
\end{array}
$$
That is, if the initial memory configurations agree on their respective low parts, then the same should hold true of the final memory configurations. A visual representation of this requirements is provided in Figure \ref{noninterfere}.

\begin{figure}
	TODO!!!
	\caption{Visual representation of the noninterference principle}
\end{figure}

Let us now revisit the example in Figure \ref{Fi:chIFS:example}. First, for the integrity vulnerability extending between {\tt getParameter} and {\tt renderToHTML}, we observe that 
the user-provided data --- belonging in the high part of the input configuration --- flows into a security-sensitive operation, and so the low part of the output configuration. Analogously, the confidential --- and thus high --- data read via the {\tt readPassword} call is released to public observers of the log, thus flowing into the low part of the output configuration.

### 2015-04-06 stranalysis.tex
missing

String analysis is a form of static analysis that focuses on string variables in specific. The goal of the analysis is to approximate the set of runtime values that such variables can potentially be assigned at different points along the program's execution. String analysis has several important applications. Examples include detection of functional defects (e.g. checking whether format constraints on string inputs are enforced properly), code comprehension (e.g. computing the set of URLs that the client side of a web application is communicating with) and optimization (e.g. to batch together SQL queries or remove redundant input checks).

Among the main applications of string analysis are security verification and runtime enforcement. This is in light of the observation that inputs from remote soruces --- e.g. HTTP parameters in the case of web applications and data arguments passed as part of inter-application communication in the Android platform --- are serialized and received as string objects. This distinguishes strings as the most important target for security analysis.

\subsection{Theoretical Challenge}

Analogously to other forms of semantic static analysis, string analysis is an undecidable problem. This is not only for 
the inherent challenges posed by the programming language (loops, branching, etc), but also because of the complexity of approximating string values. 
 follows from the fact that the family of context-free languages (CFLs) is not closed under intersection, and so modeling string operations atop a CFL approximation of string variables is decidable only for emptiness checking, but not for inclusion and equivalence checking.

### 2015-04-06 taintanalysis.tex
missing

\section{Taint Analysis}

Perhaps the most popular method to track flow of information throughout the program via program analysis is \emph{taint analysis}. The idea, stated simply, is to assign a \emph{taint tag} to every value resulting from a source statement, where the tag is merely a bit indicating that the value is security relevant. The taint tags are then propagated along the program's control-flow paths.

A main reason for the wide adoption of taint analysis as a framework for program-analysis-based information-flow security is its simplicity and scalability. Practical security products are required to scale to industry-level programs, which may consist of miillions of lines of code, and so an analysis that tracks minimal information is favorable.

At the same time, the limited expressiveness of taint analysis brings about accuracy challenges.
